---
title: "Motion-Corrected Analysis (Interactive)"
author: "Tien"
date: "2026-07-26"
output:
  html_document:
    toc: true
    toc_float: true
runtime: shiny
---

```{r setup, include=FALSE}
required_packages <- c(
  "shiny", "tidyverse", "lme4", "lmerTest", "patchwork",
  "flextable", "emmeans", "ggh4x"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install these packages before running the document: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(shiny)
library(tidyverse)
library(lme4)
library(lmerTest)
library(patchwork)
library(flextable)
library(emmeans)
library(ggh4x)
# The manuscript figures finish with a gtable edit (the divider beside the
# average column), which needs grid's grob constructors and unit arithmetic.
library(grid)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

<style>
.main-container {
  max-width: 1500px !important;
}
.shiny-plot-output {
  width: 100% !important;
}
</style>

# Choose the motion-exclusion rule

Choose the framewise-displacement (FD) cutoff and the percentage of excluded
TRs that causes a run to be removed. The models, sample summary, and plots
below update automatically. The exported CSVs are always written for both
manuscript rules (FD 0.4 mm / 10% and FD 0.5 mm / 20%) regardless of the
selection here.

```{r controls}
fluidRow(
  column(
    width = 5,
    selectInput(
      "fd_cutoff",
      "FD cutoff",
      choices = c("0.5 mm" = "0.5", "0.4 mm" = "0.4"),
      selected = "0.4"
    )
  ),
  column(
    width = 7,
    sliderInput(
      "pct_cutoff",
      "Exclude a run when percentage of excluded TRs is at least:",
      min = 0,
      max = 100,
      value = 10,
      step = 1,
      post = "%"
    )
  )
)

helpText(
  "The primary rule is 0.4 mm / 10%; the liberal rule is 0.5 mm / 20%."
)

observeEvent(input$fd_cutoff, {
  updateSliderInput(
    session,
    "pct_cutoff",
    value = if (identical(input$fd_cutoff, "0.4")) 10 else 20
  )
}, ignoreInit = TRUE)
```

# Data preparation

Only Waves 1 and 2 are analysed. Wave 3 (MIND-C) rows are dropped from the
prepared data before anything else happens, so the motion-exclusion accounting,
the demographics, and the models all describe the same Waves 1--2 sample.

```{r locate-data, include=FALSE}
# The document reads the prepared outcome files and the run-level motion
# summaries. Set MIND_DATA_DIR (or MIND_PROJECT_DIR, whose "data" subdirectory
# is used) to override the search; otherwise the repository's data/ directory is
# found relative to the working directory, which is the script/ folder when the
# document is launched with runtime: shiny.
required_data_files <- c(
  "df_mind_all.csv",
  "df_mind_all_mc04.csv",
  "motion_exclusion_by_run_5mm.csv",
  "motion_exclusion_by_run_4mm.csv"
)

candidate_data_dirs <- c(
  Sys.getenv("MIND_DATA_DIR", unset = ""),
  file.path(Sys.getenv("MIND_PROJECT_DIR", unset = ""), "data"),
  file.path(Sys.getenv("MIND_PROJECT_DIR", unset = ""), "Data", "Motion"),
  "data",
  file.path("..", "data"),
  file.path("..", "..", "data")
)
candidate_data_dirs <- candidate_data_dirs[nzchar(candidate_data_dirs)]

has_required_files <- function(path) {
  dir.exists(path) &&
    all(file.exists(file.path(path, required_data_files)))
}

data_dir <- candidate_data_dirs[vapply(
  candidate_data_dirs, has_required_files, logical(1)
)]
if (length(data_dir) == 0) {
  stop(
    "Could not find the MiND data directory. Set MIND_DATA_DIR to the folder ",
    "containing ", paste(required_data_files, collapse = ", "), "."
  )
}
data_dir <- normalizePath(data_dir[[1]], mustWork = TRUE)
project_dir <- normalizePath(file.path(data_dir, ".."), mustWork = TRUE)
output_dir <- file.path(project_dir, "output")
```

```{r load-and-prepare, include=FALSE}
# Waves 1 and 2 only. The exploratory three-wave MIND-C analysis has been
# removed from the manuscript and its supplement, so Wave 3 never enters the
# pipeline.
ANALYSIS_WAVES <- c(1, 2)
OUTCOMES <- c("Reliability", "Confusability", "Distinctiveness")
MODEL_TERMS <- c(
  "within_cAge", "between_cAge", "Sex_M1", "Education",
  "aud_mot", "aud_vis"
)
OLDER_ADULT_AGE <- 65
MAD_THRESHOLD <- 3

# The prepared outcome files hold the 2,000-vertex motion-corrected measures:
# df_mind_all.csv is the FD 0.5 mm scrubbing pipeline and df_mind_all_mc04.csv
# the FD 0.4 mm pipeline. Each is keyed to the matching run-level motion file.
outcome_files <- c(
  "0.5" = "df_mind_all.csv",
  "0.4" = "df_mind_all_mc04.csv"
)
motion_files <- c(
  "0.5" = "motion_exclusion_by_run_5mm.csv",
  "0.4" = "motion_exclusion_by_run_4mm.csv"
)

read_data_file <- function(filename) {
  readr::read_csv(
    file.path(data_dir, filename),
    num_threads = 1,
    show_col_types = FALSE
  )
}

prepare_outcome_data <- function(data) {
  data %>%
    filter(!is.na(SubNum)) %>%
    pivot_longer(
      cols = matches("^(Aud|Mot|Vis)(Distinctiveness|Reliability|Confusability)$"),
      names_to = c("Modality", ".value"),
      names_pattern = "(Aud|Mot|Vis)(.*)"
    ) %>%
    mutate(Modality = factor(Modality, levels = c("Aud", "Mot", "Vis"))) %>%
    # Wave 3 is dropped here, i.e. before the motion-exclusion accounting, so
    # the run counts reported in the manuscript are Waves 1-2 only.
    filter(Wave_Num %in% ANALYSIS_WAVES)
}

outcome_data <- lapply(outcome_files, function(filename) {
  prepare_outcome_data(read_data_file(filename))
})

motion_exclusions <- lapply(motion_files, read_data_file)

healthy_subject_ids <- bind_rows(outcome_data) %>%
  filter(Subgroup == 1) %>%
  distinct(Subject) %>%
  pull(Subject)

subject_age_groups <- bind_rows(outcome_data) %>%
  filter(Subgroup == 1) %>%
  distinct(Subject, Age_DuringParticipation_) %>%
  transmute(
    Subject,
    Age_Group = factor(
      if_else(
        Age_DuringParticipation_ < OLDER_ADULT_AGE,
        "Young adults",
        "Older adults"
      ),
      levels = c("Young adults", "Older adults")
    )
  ) %>%
  distinct(Subject, Age_Group)
```

```{r analysis-pipeline, include=FALSE}
# Everything below is written as plain functions of (fd_cutoff, pct_cutoff) so
# the reactive display and the CSV export can share one implementation.

build_run_flags <- function(fd_cutoff, pct_cutoff) {
  motion_exclusions[[fd_cutoff]] %>%
    filter(
      subject %in% healthy_subject_ids,
      session == "placebo",
      run == "001",
      experiment %in% c("auditory", "visual", "motor")
    ) %>%
    transmute(
      Subject = subject,
      Modality = recode(
        experiment,
        auditory = "Aud",
        motor = "Mot",
        visual = "Vis"
      ),
      pct_excluded,
      exclude = pct_excluded >= pct_cutoff
    )
}

build_joined_data <- function(fd_cutoff, pct_cutoff) {
  outcome_data[[fd_cutoff]] %>%
    left_join(
      build_run_flags(fd_cutoff, pct_cutoff),
      by = c("Subject", "Modality")
    )
}

build_motion_filtered_data <- function(fd_cutoff, pct_cutoff) {
  build_joined_data(fd_cutoff, pct_cutoff) %>%
    # Intended behaviour, confirmed against the reported sample sizes: a
    # session x modality with NO run-level motion record is DROPPED, not
    # retained. `!exclude` already drops NA, but the NA test is written out so
    # the intent is explicit and is not mistaken for the earlier
    # `is.na(exclude) | !exclude` behaviour, which kept unassessed rows. See
    # "Missing motion records" below for how many rows this affects.
    filter(!is.na(exclude), !exclude)
}

count_missing_motion_records <- function(fd_cutoff, pct_cutoff) {
  build_joined_data(fd_cutoff, pct_cutoff) %>%
    filter(Subgroup == 1, is.na(exclude)) %>%
    nrow()
}

add_age_decomposition <- function(data) {
  data %>%
    group_by(SubNum) %>%
    mutate(
      n_waves = n_distinct(Wave_Num),
      mean_age = if_else(
        n_waves > 1,
        mean(Age_DuringParticipation_, na.rm = TRUE),
        Age_DuringParticipation_
      ),
      within_cAge = if_else(
        n_waves > 1,
        Age_DuringParticipation_ - mean_age,
        0
      ),
      between_cAge = mean_age - OLDER_ADULT_AGE,
      aud_mot = case_when(
        Modality == "Aud" ~ -1,
        Modality == "Mot" ~ 1,
        TRUE ~ 0
      ),
      aud_vis = case_when(
        Modality == "Aud" ~ -1,
        Modality == "Vis" ~ 1,
        TRUE ~ 0
      )
    ) %>%
    ungroup()
}

build_analysis_data <- function(fd_cutoff, pct_cutoff) {
  build_motion_filtered_data(fd_cutoff, pct_cutoff) %>%
    filter(Subgroup == 1, Wave_Num %in% ANALYSIS_WAVES) %>%
    add_age_decomposition()
}

fit_model <- function(data, outcome) {
  model_data <- data %>%
    drop_na(all_of(c(outcome, MODEL_TERMS, "SubNum")))

  if (nrow(model_data) == 0) {
    stop("No complete observations are available for ", outcome)
  }

  lmer(
    reformulate(
      c(MODEL_TERMS, "(1 | SubNum)"),
      response = outcome
    ),
    data = model_data,
    REML = FALSE
  )
}

fit_outcome_models <- function(data) {
  setNames(
    lapply(OUTCOMES, function(outcome) fit_model(data, outcome)),
    OUTCOMES
  )
}

rule_text <- function(fd_cutoff, pct_cutoff) {
  paste0(
    "FD ≥ ", fd_cutoff, " mm; run excluded at ≥ ",
    pct_cutoff, "% excluded TRs"
  )
}

# The two rules the manuscript reports. `prefix` is used for rule-specific
# export filenames.
MOTION_RULES <- list(
  list(
    key = "primary",
    fd_cutoff = "0.4",
    pct_cutoff = 10,
    label = "FD 0.4 mm / 10% (primary)",
    prefix = "fd04_pct10"
  ),
  list(
    key = "liberal",
    fd_cutoff = "0.5",
    pct_cutoff = 20,
    label = "FD 0.5 mm / 20% (liberal)",
    prefix = "fd05_pct20"
  )
)

format_p <- function(p) {
  if_else(
    is.na(p),
    NA_character_,
    if_else(p < .001, "< .001", sprintf("%.3f", p))
  )
}

# The analytic data for each manuscript rule is built once here and reused by
# every static (non-reactive) section: exports, the interaction models, and the
# manuscript figures.
analysis_data_by_rule <- setNames(
  lapply(MOTION_RULES, function(rule) {
    build_analysis_data(rule$fd_cutoff, rule$pct_cutoff)
  }),
  vapply(MOTION_RULES, `[[`, character(1), "key")
)
```

```{r reactive-analysis, include=FALSE}
selected_flags <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  build_run_flags(input$fd_cutoff, input$pct_cutoff)
})

analysis_data <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  build_analysis_data(input$fd_cutoff, input$pct_cutoff)
})

missing_motion_records <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  count_missing_motion_records(input$fd_cutoff, input$pct_cutoff)
})

models <- reactive({
  fit_outcome_models(analysis_data())
})

rule_label <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  rule_text(input$fd_cutoff, input$pct_cutoff)
})
```

# Sample after exclusion

```{r sample-summary}
renderTable({
  flags <- selected_flags()
  kept <- analysis_data()

  tibble(
    `Selected rule` = rule_label(),
    `Runs assessed` = nrow(flags),
    `Runs flagged` = sum(flags$exclude, na.rm = TRUE),
    `Runs flagged (%)` = sprintf(
      "%.1f%%",
      100 * mean(flags$exclude, na.rm = TRUE)
    ),
    `Unique scans retained` = n_distinct(kept$Subject),
    `Unique participants retained` = n_distinct(kept$SubNum),
    `Modality-level observations retained` = nrow(kept),
    `Rows dropped for a missing motion record` = missing_motion_records()
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Missing motion records

`filter(!exclude)` drops any session x modality that has no run-level motion
record at all, because such a row was never assessed for motion. This is the
intended behaviour: it reproduces the sample sizes reported in the manuscript
(724 observations under the primary rule, 841 under the liberal rule). The
earlier `filter(is.na(exclude) | !exclude)` retained unassessed rows and gives
larger samples (808 and 925 respectively). The count below is the number of
healthy Waves 1--2 rows removed for a *missing* record rather than for an
actual motion exclusion; it does not depend on the FD or percentage cutoff and
should be described in the manuscript.

```{r missing-motion-records}
renderTable({
  bind_rows(lapply(MOTION_RULES, function(rule) {
    joined <- build_joined_data(rule$fd_cutoff, rule$pct_cutoff) %>%
      filter(Subgroup == 1)
    tibble(
      `Motion rule` = rule$label,
      `Healthy Waves 1-2 rows before motion filtering` = nrow(joined),
      `Rows with no motion record` = sum(is.na(joined$exclude)),
      `Rows excluded for motion` = sum(joined$exclude, na.rm = TRUE),
      `Rows retained` = sum(!is.na(joined$exclude) & !joined$exclude)
    )
  }))
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# APA-style demographics and descriptive statistics

These tables use the currently selected motion-exclusion rule. The
cross-sectional sample is defined as Wave 1. The longitudinal sample requires
retained observations at both Waves 1 and 2; Wave 1 demographics are reported
so each longitudinal participant is counted once. Young adults are below age
65 and older adults are age 65 or above. In Table 2, longitudinal descriptive
statistics average each participant's two retained waves within modality
before calculating the group mean and standard deviation.

```{r apa-table-data, include=FALSE}
crosssec_data <- reactive({
  analysis_data() %>%
    filter(Wave_Num == 1) %>%
    mutate(
      Age_Group = factor(
        if_else(
          Age_DuringParticipation_ < OLDER_ADULT_AGE,
          "Young adults",
          "Older adults"
        ),
        levels = c("Young adults", "Older adults")
      )
    )
})

demographics_summary <- reactive({
  crosssec_participants <- crosssec_data() %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Cross-sectional")

  longitudinal_ids <- analysis_data() %>%
    distinct(SubNum, Wave_Num) %>%
    count(SubNum, name = "n_retained_waves") %>%
    filter(n_retained_waves >= 2) %>%
    pull(SubNum)

  longitudinal_participants <- crosssec_data() %>%
    filter(SubNum %in% longitudinal_ids) %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_participants, longitudinal_participants) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group) %>%
    summarise(
      n = n(),
      Age_M = mean(Age_DuringParticipation_, na.rm = TRUE),
      Age_SD = sd(Age_DuringParticipation_, na.rm = TRUE),
      Education_M = mean(Education, na.rm = TRUE),
      Education_SD = sd(Education, na.rm = TRUE),
      Male_n = sum(Sex_M1 == 1, na.rm = TRUE),
      Female_n = sum(Sex_M1 == 2, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      `Age, M (SD)` = sprintf("%.2f (%.2f)", Age_M, Age_SD),
      `Education, M (SD)` = sprintf(
        "%.2f (%.2f)",
        Education_M,
        Education_SD
      ),
      `Male, n (%)` = sprintf("%d (%.1f%%)", Male_n, 100 * Male_n / n),
      `Female, n (%)` = sprintf(
        "%d (%.1f%%)",
        Female_n,
        100 * Female_n / n
      )
    ) %>%
    select(
      Sample,
      `Age group` = Age_Group,
      N = n,
      `Age, M (SD)`,
      `Education, M (SD)`,
      `Male, n (%)`,
      `Female, n (%)`
    )
})

descriptives_summary <- reactive({
  crosssec_descriptives <- crosssec_data() %>%
    group_by(SubNum, Age_Group, Modality) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(Sample = "Cross-sectional")

  baseline_age_groups <- crosssec_data() %>%
    distinct(SubNum, Age_Group)

  longitudinal_descriptives <- analysis_data() %>%
    group_by(SubNum, Modality) %>%
    filter(n_distinct(Wave_Num) >= 2) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    inner_join(baseline_age_groups, by = "SubNum") %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_descriptives, longitudinal_descriptives) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group, Modality) %>%
    summarise(
      Reliability_M = mean(Reliability, na.rm = TRUE),
      Reliability_SD = sd(Reliability, na.rm = TRUE),
      Confusability_M = mean(Confusability, na.rm = TRUE),
      Confusability_SD = sd(Confusability, na.rm = TRUE),
      Distinctiveness_M = mean(Distinctiveness, na.rm = TRUE),
      Distinctiveness_SD = sd(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      Modality = recode(
        as.character(Modality),
        Aud = "Auditory",
        Mot = "Motor",
        Vis = "Visual"
      )
    ) %>%
    arrange(
      Sample,
      Age_Group,
      factor(Modality, levels = c("Auditory", "Motor", "Visual"))
    )
})

style_apa_table <- function(ft) {
  ft %>%
    theme_booktabs() %>%
    font(fontname = "Times New Roman", part = "all") %>%
    fontsize(size = 11, part = "all") %>%
    bold(part = "header") %>%
    align(align = "center", part = "all") %>%
    align(j = 1, align = "left", part = "all") %>%
    padding(padding = 4, part = "all") %>%
    autofit()
}

demographics_flextable <- reactive({
  flextable(demographics_summary()) %>%
    merge_v(j = "Sample") %>%
    valign(j = "Sample", valign = "top") %>%
    style_apa_table()
})

descriptives_flextable <- reactive({
  flextable(descriptives_summary()) %>%
    colformat_double(
      j = c(
        "Reliability_M", "Reliability_SD",
        "Confusability_M", "Confusability_SD",
        "Distinctiveness_M", "Distinctiveness_SD"
      ),
      digits = 3,
      na_str = "—"
    ) %>%
    set_header_labels(
      Sample = "Sample",
      Age_Group = "Age group",
      Modality = "Modality",
      Reliability_M = "M",
      Reliability_SD = "SD",
      Confusability_M = "M",
      Confusability_SD = "SD",
      Distinctiveness_M = "M",
      Distinctiveness_SD = "SD"
    ) %>%
    add_header_row(
      values = c(
        "Sample", "Age group", "Modality", "Reliability",
        "Confusability", "Distinctiveness"
      ),
      colwidths = c(1, 1, 1, 2, 2, 2)
    ) %>%
    merge_v(j = "Sample") %>%
    merge_v(
      j = c("Sample", "Age_Group"),
      target = "Age_Group",
      combine = TRUE
    ) %>%
    valign(
      j = c("Sample", "Age_Group"),
      valign = "top"
    ) %>%
    style_apa_table()
})

apa_table_html <- function(number, title, note, ft) {
  htmltools::tagList(
    htmltools::tags$p(
      htmltools::tags$strong(number),
      style = "font-family: 'Times New Roman'; margin-bottom: 0;"
    ),
    htmltools::tags$p(
      htmltools::tags$em(title),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 1.05em; margin-top: 0;"
      )
    ),
    flextable::htmltools_value(ft),
    htmltools::tags$p(
      htmltools::tags$em("Note. "),
      paste0(note, " Selected rule: ", rule_label(), "."),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 0.95em; margin-top: 8px;"
      )
    )
  )
}
```

## Table 1

```{r demographics-table}
renderUI({
  apa_table_html(
    "Table 1",
    "Cross-Sectional and Longitudinal Analysis Demographics Summary",
    paste(
      "Values are presented as M (SD) unless otherwise indicated.",
      "The cross-sectional sample includes retained Wave 1 participants.",
      "The longitudinal sample includes participants with retained",
      "observations at Waves 1 and 2; their Wave 1 demographics are shown.",
      "Participants may therefore appear in both samples."
    ),
    demographics_flextable()
  )
})

downloadButton("download_demographics", "Download Table 1 (.docx)")

output$download_demographics <- downloadHandler(
  filename = function() "table1_demographics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 1. Cross-Sectional and Longitudinal Demographics" =
        demographics_flextable(),
      path = file
    )
  }
)
```

## Table 2

```{r descriptives-table}
renderUI({
  apa_table_html(
    "Table 2",
    "Cross-Sectional and Longitudinal Analysis Descriptive Statistics",
    paste(
      "Values are means and standard deviations for reliability",
      "(within-category similarity), confusability (between-category",
      "similarity), and distinctiveness by sample, age group, and modality.",
      "Longitudinal values are participant-level averages across two",
      "retained waves within each modality."
    ),
    descriptives_flextable()
  )
})

downloadButton("download_descriptives", "Download Table 2 (.docx)")

output$download_descriptives <- downloadHandler(
  filename = function() "table2_descriptive_statistics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 2. Cross-Sectional and Longitudinal Descriptives" =
        descriptives_flextable(),
      path = file
    )
  }
)
```

# Mixed-effects models

The fixed-effects specification is unchanged from the original document:
within-person age, between-person age, sex, education, and the two
effect-coded modality contrasts, with a participant random intercept.

```{r model-choice}
selectInput(
  "outcome",
  "Outcome shown below",
  choices = c("Reliability", "Confusability", "Distinctiveness"),
  selected = "Reliability"
)
```

## Model summary

```{r model-summary}
renderPrint({
  cat(rule_label(), "\n\n")
  print(summary(models()[[input$outcome]]))
})
```

## Type III tests

Fixed effects are tested with Satterthwaite-approximated denominator degrees of
freedom, consistent with the interaction tests below.

```{r model-anova}
renderPrint({
  cat(rule_label(), "\n\n")
  print(stats::anova(
    models()[[input$outcome]],
    type = 3,
    ddf = "Satterthwaite"
  ))
})
```

# Modality-specific age effects

Each outcome is refit with the two age terms crossed with modality:

```
outcome ~ (between_cAge + within_cAge) * Modality +
          Sex_M1 + Education + (1 | SubNum)
```

`Modality` enters as a factor (Auditory, Visual, Motor) rather than the manual
`aud_mot` / `aud_vis` contrasts, and `contr.sum` is in effect for these fits so
the Type III tests are valid. The joint interaction tests are the 2-df
Satterthwaite *F* tests for `between_cAge:Modality` and `within_cAge:Modality`
from `anova()`.

Simple slopes are probed with `emmeans::emtrends()` on the interaction model —
not with stratified per-modality models — and pairwise slope differences use
Holm adjustment across the three modality pairs. All degrees of freedom are
Satterthwaite.

Everything is computed for both age terms and all three outcomes so the exports
are complete. The `interaction_significant` column marks the joint *F* test at
*p* < .05, which is the gate the manuscript's pre-registered rule uses: only a
significant interaction is decomposed and interpreted.

```{r interaction-models, include=FALSE}
INTERACTION_MODALITY_LEVELS <- c("Auditory", "Visual", "Motor")
INTERACTION_CONTRASTS <- c("contr.sum", "contr.poly")
AGE_TERMS <- c("between_cAge", "within_cAge")

prepare_interaction_data <- function(data) {
  data %>%
    mutate(
      Modality = factor(
        recode(
          as.character(Modality),
          Aud = "Auditory",
          Vis = "Visual",
          Mot = "Motor"
        ),
        levels = INTERACTION_MODALITY_LEVELS
      )
    )
}

fit_interaction_model <- function(data, outcome) {
  # contr.sum is required for a valid Type III test of a factor that appears in
  # an interaction; the previous setting is restored on exit so the main models
  # and their manual numeric contrasts are untouched.
  previous_contrasts <- options(contrasts = INTERACTION_CONTRASTS)
  on.exit(options(previous_contrasts), add = TRUE)

  model_data <- prepare_interaction_data(data) %>%
    drop_na(all_of(c(
      outcome, "between_cAge", "within_cAge", "Sex_M1", "Education", "SubNum"
    )))

  lmer(
    reformulate(
      c(
        "(between_cAge + within_cAge) * Modality",
        "Sex_M1", "Education", "(1 | SubNum)"
      ),
      response = outcome
    ),
    data = model_data,
    REML = FALSE
  )
}

fit_interaction_models <- function(data) {
  setNames(
    lapply(OUTCOMES, function(outcome) fit_interaction_model(data, outcome)),
    OUTCOMES
  )
}

interaction_tests_table <- function(interaction_models) {
  map_dfr(OUTCOMES, function(outcome) {
    # lmerTest's anova() is Type III with Satterthwaite denominator df.
    anova_table <- anova(interaction_models[[outcome]])

    map_dfr(AGE_TERMS, function(age_term) {
      row <- paste0(age_term, ":Modality")
      tibble(
        outcome = outcome,
        age_term = age_term,
        F = anova_table[row, "F value"],
        df1 = anova_table[row, "NumDF"],
        df2 = anova_table[row, "DenDF"],
        p = anova_table[row, "Pr(>F)"]
      )
    })
  })
}

modality_trends <- function(model, age_term) {
  # lmer.df is always passed explicitly so emmeans never falls back to
  # Kenward-Roger for these models.
  emmeans::emtrends(
    model,
    ~ Modality,
    var = age_term,
    lmer.df = "satterthwaite",
    infer = c(TRUE, TRUE)
  )
}

simple_slopes_table <- function(interaction_models, tests) {
  gate <- tests %>%
    transmute(outcome, age_term, interaction_significant = p < .05)

  map_dfr(OUTCOMES, function(outcome) {
    map_dfr(AGE_TERMS, function(age_term) {
      summary(modality_trends(interaction_models[[outcome]], age_term)) %>%
        as_tibble() %>%
        transmute(
          outcome = outcome,
          age_term = age_term,
          modality = as.character(Modality),
          slope = .data[[paste0(age_term, ".trend")]],
          SE,
          df,
          lower.CL,
          upper.CL,
          t = t.ratio,
          p = p.value
        )
    })
  }) %>%
    left_join(gate, by = c("outcome", "age_term"))
}

slope_contrasts_table <- function(interaction_models) {
  map_dfr(OUTCOMES, function(outcome) {
    map_dfr(AGE_TERMS, function(age_term) {
      trends <- modality_trends(interaction_models[[outcome]], age_term)
      summary(pairs(trends, adjust = "holm")) %>%
        as_tibble() %>%
        transmute(
          outcome = outcome,
          age_term = age_term,
          contrast = as.character(contrast),
          estimate,
          SE,
          df,
          t = t.ratio,
          p_holm = p.value
        )
    })
  })
}

compute_interaction_results <- function(data) {
  interaction_models <- fit_interaction_models(data)
  tests <- interaction_tests_table(interaction_models)

  list(
    models = interaction_models,
    tests = tests,
    slopes = simple_slopes_table(interaction_models, tests),
    contrasts = slope_contrasts_table(interaction_models)
  )
}

interaction_results_by_rule <- lapply(
  analysis_data_by_rule,
  compute_interaction_results
)

interaction_results <- reactive({
  compute_interaction_results(analysis_data())
})

format_interaction_tests <- function(tests) {
  tests %>%
    transmute(
      Outcome = outcome,
      `Age term` = age_term,
      F = sprintf("%.2f", F),
      df1,
      df2 = sprintf("%.1f", df2),
      p = format_p(p),
      `Interaction significant` = if_else(p < .05, "Yes", "No")
    )
}

format_simple_slopes <- function(slopes) {
  slopes %>%
    transmute(
      Outcome = outcome,
      `Age term` = age_term,
      Modality = modality,
      Slope = sprintf("%.5f", slope),
      SE = sprintf("%.5f", SE),
      df = sprintf("%.1f", df),
      `95% CI` = sprintf("[%.5f, %.5f]", lower.CL, upper.CL),
      t = sprintf("%.2f", t),
      p = format_p(p),
      `Interaction significant` = if_else(interaction_significant, "Yes", "No")
    )
}

format_slope_contrasts <- function(contrasts) {
  contrasts %>%
    transmute(
      Outcome = outcome,
      `Age term` = age_term,
      Contrast = contrast,
      Estimate = sprintf("%.5f", estimate),
      SE = sprintf("%.5f", SE),
      df = sprintf("%.1f", df),
      t = sprintf("%.2f", t),
      `p (Holm)` = format_p(p_holm)
    )
}
```

## Joint interaction tests

```{r interaction-tests-table}
renderTable({
  format_interaction_tests(interaction_results()$tests)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Simple slopes by modality

```{r interaction-slopes-table}
renderTable({
  format_simple_slopes(interaction_results()$slopes)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Pairwise slope differences (Holm)

```{r interaction-contrasts-table}
renderTable({
  format_slope_contrasts(interaction_results()$contrasts)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# Contextual effect: do the within- and between-person age slopes differ?

The main models estimate a within-person slope (`within_cAge`) and a
between-person slope (`between_cAge`) separately, but neither *t* test asks
whether the two differ from each other. That difference, `b_between -
b_within`, is the contextual effect of Hoffman and Stawski (2009); a
contextual effect indistinguishable from zero means one common age slope
describes both the cross-sectional and the longitudinal change.

It is computed here two ways, and the two must agree:

1. **Reparameterisation.** `age_session_c = within_cAge + between_cAge` is
   session age centred at `r OLDER_ADULT_AGE`, so replacing the pair
   (`within_cAge`, `between_cAge`) with (`age_session_c`, `between_cAge`)
   spans the same column space and is the same model refit in new
   coordinates. In those coordinates the slope on `age_session_c` *is* the
   within-person slope and the coefficient on `between_cAge` *is* the
   contextual effect, so its Satterthwaite *t* test is the test of
   H0: `b_between = b_within`.
2. **Contrast on the original model.** `lmerTest::contest1D()` with
   *L* = -1 on `within_cAge`, +1 on `between_cAge`, zero elsewhere.

Both routes use the estimator settings of the main models (`REML = FALSE`,
Satterthwaite df) and the same complete-case rows. The chunk below asserts the
agreement rather than trusting it: the reparameterised fit must reproduce the
original within-person slope and log-likelihood, and the two routes must return
the same estimate, SE, df, *t*, and *p*.

```{r contextual-effects, include=FALSE}
# Estimates, SEs, t and p agree to floating-point noise. Satterthwaite df are
# the exception: they come from numerical derivatives of the profiled
# likelihood, so the two parameterisations agree only to ~1e-5 and get their
# own tolerance.
CONTEXTUAL_TOLERANCE <- 1e-8
CONTEXTUAL_DF_TOLERANCE <- 1e-3

# Same rows, same estimator, same random effect as fit_model(); only the two
# age columns are re-expressed.
fit_contextual_model <- function(data, outcome) {
  model_data <- data %>%
    drop_na(all_of(c(outcome, MODEL_TERMS, "SubNum"))) %>%
    mutate(age_session_c = within_cAge + between_cAge)

  lmer(
    reformulate(
      c("age_session_c", setdiff(MODEL_TERMS, "within_cAge"), "(1 | SubNum)"),
      response = outcome
    ),
    data = model_data,
    REML = FALSE
  )
}

contextual_contrast <- function(model) {
  L <- setNames(rep(0, length(fixef(model))), names(fixef(model)))
  L[["within_cAge"]] <- -1
  L[["between_cAge"]] <- 1
  lmerTest::contest1D(model, L, ddf = "Satterthwaite")
}

contextual_effects_table <- function(data, rule_label_text) {
  map_dfr(OUTCOMES, function(outcome) {
    model <- fit_model(data, outcome)
    reparameterised <- fit_contextual_model(data, outcome)

    original <- summary(model)$coefficients
    repar <- summary(reparameterised)$coefficients
    contrast <- contextual_contrast(model)

    # A reparameterisation, not a different model: the slope on session age is
    # the original within-person slope and the fit is identical.
    stopifnot(
      abs(repar["age_session_c", "Estimate"] -
        original["within_cAge", "Estimate"]) < CONTEXTUAL_TOLERANCE,
      abs(as.numeric(logLik(reparameterised)) - as.numeric(logLik(model))) <
        CONTEXTUAL_TOLERANCE
    )

    result <- tibble(
      motion_rule = rule_label_text,
      outcome = outcome,
      estimate = repar["between_cAge", "Estimate"],
      se = repar["between_cAge", "Std. Error"],
      df = repar["between_cAge", "df"],
      t = repar["between_cAge", "t value"],
      p = repar["between_cAge", "Pr(>|t|)"]
    )

    # ... and the contrast on the original model must land in the same place.
    stopifnot(
      abs(result$estimate - contrast[["Estimate"]]) < CONTEXTUAL_TOLERANCE,
      abs(result$se - contrast[["Std. Error"]]) < CONTEXTUAL_TOLERANCE,
      abs(result$t - contrast[["t value"]]) < CONTEXTUAL_TOLERANCE,
      abs(result$p - contrast[["Pr(>|t|)"]]) < CONTEXTUAL_TOLERANCE,
      abs(result$df - contrast[["df"]]) < CONTEXTUAL_DF_TOLERANCE
    )

    result
  })
}

format_contextual_effects <- function(table) {
  table %>%
    transmute(
      Outcome = outcome,
      `Contextual effect` = sprintf("%.5f", estimate),
      SE = sprintf("%.5f", se),
      df = sprintf("%.1f", df),
      t = sprintf("%.2f", t),
      p = format_p(p),
      `Slopes differ` = if_else(p < .05, "Yes", "No")
    )
}

contextual_effects_by_rule <- setNames(
  lapply(MOTION_RULES, function(rule) {
    contextual_effects_table(analysis_data_by_rule[[rule$key]], rule$label)
  }),
  vapply(MOTION_RULES, `[[`, character(1), "key")
)
```

## Contextual effects for both manuscript rules

These do not depend on the controls above; they are the numbers written to
`contextual_effects.csv`.

```{r contextual-effects-table}
renderUI({
  htmltools::tagList(lapply(MOTION_RULES, function(rule) {
    htmltools::tagList(
      htmltools::tags$h4(rule$label),
      htmltools::HTML(
        knitr::kable(
          format_contextual_effects(contextual_effects_by_rule[[rule$key]]),
          format = "html",
          row.names = FALSE,
          table.attr = "class='table table-striped table-bordered'"
        )
      )
    )
  }))
})
```

# Outlier sensitivity (MAD x 3)

Outliers are flagged with a median absolute deviation criterion applied to each
outcome **separately within each modality**, because the auditory, motor, and
visual distributions differ enough that pooling them inflates the MAD and
flags almost nothing. An observation is flagged when
`abs(x - median(x)) > 3 * mad(x)`; R's `mad()` already applies the 1.4826
consistency constant.

Two exclusion variants are then compared with the full sample. A row is treated
as flagged if it is an outlier on *any* of the three outcomes, so all three
models are refit on the same subsample and the estimates stay comparable
across outcomes.

- **Drop flagged observations** removes only the flagged session x modality rows.
- **Drop flagged participants** removes every row belonging to a participant
  with at least one flagged observation.

The model specification is unchanged: `outcome ~ within_cAge + between_cAge +
Sex_M1 + Education + aud_mot + aud_vis + (1 | SubNum)`, `REML = FALSE`, with
Satterthwaite *p* values from `lmerTest`.

```{r outlier-sensitivity-helpers, include=FALSE}
add_mad_outlier_flags <- function(data) {
  data %>%
    group_by(Modality) %>%
    mutate(
      across(
        all_of(OUTCOMES),
        ~ !is.na(.x) &
          abs(.x - median(.x, na.rm = TRUE)) >
            MAD_THRESHOLD * mad(.x, na.rm = TRUE),
        .names = "outlier_{.col}"
      )
    ) %>%
    ungroup() %>%
    mutate(
      outlier_any = if_any(starts_with("outlier_"))
    )
}

mad_outlier_counts <- function(data) {
  flagged <- add_mad_outlier_flags(data)

  by_outcome <- map_dfr(OUTCOMES, function(outcome) {
    indicator <- flagged[[paste0("outlier_", outcome)]]
    tibble(
      Outcome = outcome,
      `Observations flagged` = sum(indicator),
      `Observations flagged (%)` = 100 * mean(indicator),
      `Participants affected` = n_distinct(flagged$SubNum[indicator])
    )
  })

  bind_rows(
    by_outcome,
    tibble(
      Outcome = "Any outcome",
      `Observations flagged` = sum(flagged$outlier_any),
      `Observations flagged (%)` = 100 * mean(flagged$outlier_any),
      `Participants affected` =
        n_distinct(flagged$SubNum[flagged$outlier_any])
    )
  )
}

build_outlier_variants <- function(data) {
  flagged <- add_mad_outlier_flags(data)
  affected_participants <- flagged %>%
    filter(outlier_any) %>%
    distinct(SubNum) %>%
    pull(SubNum)

  list(
    `Full sample` = flagged,
    `Drop flagged observations` = flagged %>% filter(!outlier_any),
    `Drop flagged participants` = flagged %>%
      filter(!SubNum %in% affected_participants)
  )
}

outlier_sensitivity_table <- function(data, rule_label_text = NA_character_) {
  variants <- build_outlier_variants(data)

  map_dfr(OUTCOMES, function(outcome) {
    map_dfr(names(variants), function(variant) {
      model <- fit_model(variants[[variant]], outcome)
      model_frame <- model@frame
      coefficients <- summary(model)$coefficients

      tibble(
        `Motion rule` = rule_label_text,
        Outcome = outcome,
        `Exclusion variant` = variant,
        `N observations` = nrow(model_frame),
        `N participants` = n_distinct(model_frame$SubNum),
        age_between_b = coefficients["between_cAge", "Estimate"],
        age_between_SE = coefficients["between_cAge", "Std. Error"],
        age_between_df = coefficients["between_cAge", "df"],
        age_between_p = coefficients["between_cAge", "Pr(>|t|)"],
        age_within_b = coefficients["within_cAge", "Estimate"],
        age_within_SE = coefficients["within_cAge", "Std. Error"],
        age_within_df = coefficients["within_cAge", "df"],
        age_within_p = coefficients["within_cAge", "Pr(>|t|)"]
      )
    })
  })
}

format_outlier_sensitivity <- function(table) {
  table %>%
    mutate(
      across(
        c(age_between_b, age_between_SE, age_within_b, age_within_SE),
        ~ sprintf("%.5f", .x)
      ),
      across(c(age_between_df, age_within_df), ~ sprintf("%.1f", .x)),
      across(c(age_between_p, age_within_p), format_p)
    ) %>%
    select(-any_of("Motion rule")) %>%
    rename(
      `age_between b` = age_between_b,
      `age_between SE` = age_between_SE,
      `age_between df` = age_between_df,
      `age_between p` = age_between_p,
      `age_within b` = age_within_b,
      `age_within SE` = age_within_SE,
      `age_within df` = age_within_df,
      `age_within p` = age_within_p
    )
}

outlier_sensitivity_by_rule <- setNames(
  lapply(MOTION_RULES, function(rule) {
    outlier_sensitivity_table(
      analysis_data_by_rule[[rule$key]],
      rule$label
    )
  }),
  vapply(MOTION_RULES, `[[`, character(1), "key")
)
```

## Flagged observations under the selected rule

```{r outlier-counts}
renderPrint({
  data <- analysis_data()
  counts <- mad_outlier_counts(data)
  cat(rule_label(), "\n")
  cat("Analytic observations:", nrow(data), "\n")
  cat("Participants:", n_distinct(data$SubNum), "\n\n")
  for (i in seq_len(nrow(counts))) {
    cat(sprintf(
      "%-16s %3d observations flagged (%.1f%%) across %d participants\n",
      counts$Outcome[[i]],
      counts$`Observations flagged`[[i]],
      counts$`Observations flagged (%)`[[i]],
      counts$`Participants affected`[[i]]
    ))
  }
})
```

## Comparison table under the selected rule

```{r outlier-sensitivity-table}
renderTable({
  format_outlier_sensitivity(outlier_sensitivity_table(analysis_data()))
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Comparison tables for both manuscript rules

These two tables do not depend on the controls above; they are the numbers
written to `outlier_sensitivity.csv`.

```{r outlier-sensitivity-both-rules}
renderUI({
  htmltools::tagList(lapply(MOTION_RULES, function(rule) {
    table <- outlier_sensitivity_by_rule[[rule$key]]
    htmltools::tagList(
      htmltools::tags$h4(rule$label),
      htmltools::HTML(
        knitr::kable(
          format_outlier_sensitivity(table),
          format = "html",
          row.names = FALSE,
          table.attr = "class='table table-striped table-bordered'"
        )
      )
    )
  }))
})
```

# Exported numbers for the manuscript

```{r exports, include=FALSE}
sample_definitions <- function(data) {
  returner_ids <- data %>%
    distinct(SubNum, Wave_Num) %>%
    count(SubNum, name = "n_retained_waves") %>%
    filter(n_retained_waves >= 2) %>%
    pull(SubNum)

  young_wave1 <- data %>%
    filter(Wave_Num == 1, Age_DuringParticipation_ < OLDER_ADULT_AGE)
  older_wave1 <- data %>%
    filter(Wave_Num == 1, Age_DuringParticipation_ >= OLDER_ADULT_AGE)

  list(
    `Young adults, Wave 1` = young_wave1,
    `Older adults, Wave 1` = older_wave1,
    `Older adults, Wave 2 (returners)` = data %>%
      filter(
        Wave_Num == 2,
        Age_DuringParticipation_ >= OLDER_ADULT_AGE,
        SubNum %in% returner_ids
      ),
    `Older adults, Wave 1 (Wave 2 returners)` = older_wave1 %>%
      filter(SubNum %in% returner_ids)
  )
}

demographics_export <- function(data, rule_label_text) {
  samples <- sample_definitions(data)

  map_dfr(names(samples), function(sample_name) {
    participants <- samples[[sample_name]] %>%
      distinct(SubNum, .keep_all = TRUE)

    tibble(
      motion_rule = rule_label_text,
      sample = sample_name,
      n = nrow(participants),
      age_m = mean(participants$Age_DuringParticipation_, na.rm = TRUE),
      age_sd = sd(participants$Age_DuringParticipation_, na.rm = TRUE),
      age_min = min(participants$Age_DuringParticipation_, na.rm = TRUE),
      age_max = max(participants$Age_DuringParticipation_, na.rm = TRUE),
      education_m = mean(participants$Education, na.rm = TRUE),
      education_sd = sd(participants$Education, na.rm = TRUE),
      n_men = sum(participants$Sex_M1 == 1, na.rm = TRUE),
      n_women = sum(participants$Sex_M1 == 2, na.rm = TRUE)
    )
  })
}

descriptives_export <- function(data, rule_label_text) {
  samples <- sample_definitions(data)

  map_dfr(names(samples), function(sample_name) {
    samples[[sample_name]] %>%
      mutate(
        modality = recode(
          as.character(Modality),
          Aud = "Auditory",
          Mot = "Motor",
          Vis = "Visual"
        )
      ) %>%
      group_by(modality) %>%
      summarise(
        n = sum(!is.na(Reliability)),
        reliability_m = mean(Reliability, na.rm = TRUE),
        reliability_sd = sd(Reliability, na.rm = TRUE),
        confusability_m = mean(Confusability, na.rm = TRUE),
        confusability_sd = sd(Confusability, na.rm = TRUE),
        distinctiveness_m = mean(Distinctiveness, na.rm = TRUE),
        distinctiveness_sd = sd(Distinctiveness, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        motion_rule = rule_label_text,
        sample = sample_name,
        .before = 1
      )
  })
}

motion_accounting_export <- function() {
  map_dfr(MOTION_RULES, function(rule) {
    flags <- build_run_flags(rule$fd_cutoff, rule$pct_cutoff) %>%
      left_join(subject_age_groups, by = "Subject") %>%
      mutate(
        age_group = as.character(Age_Group),
        modality = recode(
          as.character(Modality),
          Aud = "Auditory",
          Mot = "Motor",
          Vis = "Visual"
        )
      )

    summarise_flags <- function(grouped) {
      grouped %>%
        summarise(
          runs_assessed = n(),
          runs_excluded = sum(exclude, na.rm = TRUE),
          pct_runs_excluded = 100 * mean(exclude, na.rm = TRUE),
          mean_pct_volumes_scrubbed = mean(pct_excluded, na.rm = TRUE),
          .groups = "drop"
        )
    }

    bind_rows(
      summarise_flags(flags %>% group_by(age_group, modality)),
      summarise_flags(flags %>% group_by(age_group)) %>%
        mutate(modality = "All modalities"),
      summarise_flags(flags %>% group_by(modality)) %>%
        mutate(age_group = "All participants"),
      summarise_flags(flags %>% group_by()) %>%
        mutate(age_group = "All participants", modality = "All modalities")
    ) %>%
      mutate(motion_rule = rule$label, .before = 1) %>%
      select(
        motion_rule, age_group, modality, runs_assessed, runs_excluded,
        pct_runs_excluded, mean_pct_volumes_scrubbed
      )
  })
}

model_estimates_export <- function() {
  map_dfr(MOTION_RULES, function(rule) {
    data <- analysis_data_by_rule[[rule$key]]
    outcome_models <- fit_outcome_models(data)

    map_dfr(OUTCOMES, function(outcome) {
      model <- outcome_models[[outcome]]
      summary(model)$coefficients %>%
        as.data.frame() %>%
        rownames_to_column("term") %>%
        as_tibble() %>%
        transmute(
          motion_rule = rule$label,
          outcome = outcome,
          term,
          estimate = Estimate,
          se = `Std. Error`,
          df,
          t = `t value`,
          p = `Pr(>|t|)`,
          n_observations = nrow(model@frame),
          n_participants = n_distinct(model@frame$SubNum)
        )
    })
  })
}

export_tables <- list()
for (rule in MOTION_RULES) {
  rule_data <- analysis_data_by_rule[[rule$key]]
  rule_interaction <- interaction_results_by_rule[[rule$key]]

  export_tables[[paste0(rule$prefix, "_demographics.csv")]] <-
    demographics_export(rule_data, rule$label)
  export_tables[[paste0(rule$prefix, "_descriptives.csv")]] <-
    descriptives_export(rule_data, rule$label)
  export_tables[[paste0(rule$prefix, "_interaction_tests.csv")]] <-
    rule_interaction$tests %>% mutate(motion_rule = rule$label, .before = 1)
  export_tables[[paste0(rule$prefix, "_simple_slopes.csv")]] <-
    rule_interaction$slopes %>% mutate(motion_rule = rule$label, .before = 1)
  export_tables[[paste0(rule$prefix, "_slope_contrasts.csv")]] <-
    rule_interaction$contrasts %>% mutate(motion_rule = rule$label, .before = 1)
}
export_tables[["motion_accounting.csv"]] <- motion_accounting_export()
export_tables[["model_estimates.csv"]] <- model_estimates_export()
export_tables[["outlier_sensitivity.csv"]] <-
  bind_rows(outlier_sensitivity_by_rule)
export_tables[["contextual_effects.csv"]] <-
  bind_rows(contextual_effects_by_rule)

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
for (filename in names(export_tables)) {
  readr::write_csv(
    export_tables[[filename]],
    file.path(output_dir, filename)
  )
}
```

The tables the manuscript needs are written to `output/` when this document
starts. `demographics.csv` and `descriptives.csv` are rule-specific and are
therefore prefixed with the motion rule; the remaining files carry a
`motion_rule` column and cover both rules in one file.

```{r export-summary}
renderTable({
  tibble(
    File = file.path("output", names(export_tables)),
    Rows = vapply(export_tables, nrow, integer(1)),
    Columns = vapply(export_tables, ncol, integer(1))
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# Manuscript figures

Figures 2 and 3 are static, publication-ready files built from the primary
motion rule; the same figures under the liberal rule are written to
`output/figures/supplement/` so they can be used as supplementary figures
without re-running anything. This section does not depend on the Shiny
controls.

Both figures are one 3 x 4 grid built by the same function, so they cannot
drift apart. Measure is the row throughout — reliability, confusability,
distinctiveness — and the four columns are the modality average, labelled
*All modalities*, followed by auditory, visual, and motor. The average is the
leftmost column of the same grid rather than a separate panel stacked above it,
and it is marked as the primary result three ways: an extra panel gap of
whitespace after it, a bold strip label, and a thin grey rule drawn in that
gap, which survives greyscale printing. All four panels keep the same white
background, and nothing else about the column differs — point size, alpha, line
width, ribbon opacity, and colour are identical across all four columns, so the
panels in a row stay directly comparable.

The grid uses `scales = "free_y"`, which frees the y scale by row and shares it
within the row: all four columns of a row sit on one scale, drawn once at the
left, so the average and the three modalities are directly comparable.
`scales = "free"` would give each of the twelve panels its own axis and destroy
that comparison, so it is not used. `switch = "y"` puts the measure strip where
the y title would otherwise be.

In Figure 2 the x-axis is age. The average column shows the pooled
between-person age effect from the main (no-interaction) model with a 95% CI
ribbon, over participant-level values averaged across modalities and waves, so
each participant appears once — session x modality points are unreadable once
pooled. The three modality columns show the same effect resolved by modality
using the slopes implied by the interaction model, so the lines correspond
exactly to the emmeans simple slopes reported above; there the points are the
individual session x modality observations. No one was scanned between the
young and older cohorts, so every model line is solid where there are data and
dashed across the unsampled 30--60 region, with the ribbon continuous
throughout.

Figure 3 has the same 3 x 4 structure for the within-person effect, on an
x-axis of years from each participant's own first retained wave rather than
chronological age, so every trajectory starts at zero. Only returning
participants appear. The average column is the pooled within-person slope from
the main model with a 95% CI ribbon, over session-level values averaged across
modalities; the modality columns resolve it by modality using the interaction
model, matching the emmeans `within_cAge` simple slopes. Their ribbons overlap
heavily across modalities, which is the honest display of a within-person x
modality interaction that is non-significant for all three outcomes. The grey
lines join each participant's own two sessions.

Both figures share one aesthetic mapping, defined once and reused by every
panel. Colour encodes the measure and nothing else (Okabe-Ito blue, vermillion,
and green for reliability, confusability, and distinctiveness) and is applied
to points, model lines, and CI ribbons. Shape encodes the column and nothing
else — open circle, triangle, and square for auditory, visual, and motor, and
a neutral open diamond for the average. Neither figure carries a legend: every
panel is named by its own strips, and a shape legend would list the average
alongside the three modalities and imply it is a fourth one. The grey two-wave
segments in Figure 3 are the raw data layer and are deliberately left
uncoloured so the model line stays visible on top of them.

Titles, subtitles, and in-panel statistics are deliberately absent: panel
identity comes from the strip labels, and the statistics belong in Table 3 and
the manuscript text.

```{r manuscript-figure-helpers, include=FALSE}
# ---------------------------------------------------------------------------
# Shared aesthetic mapping for every manuscript figure.
#
# Colour encodes the measure and nothing else. Shape encodes the modality
# column and nothing else. Both scales are defined once here and reused by
# every panel of every figure; no figure defines its own palette.
# ---------------------------------------------------------------------------
# The row order and the column order of both figures. The average column is the
# leftmost column of the same grid, not a separate panel stacked above it. It
# keeps the short level name "Average" in the data - the checks and the filters
# key on it - and is labelled "All modalities" on the figure, which describes
# the aggregate rather than suggesting a location.
MEASURES <- OUTCOMES
AVERAGE_COLUMN <- "Average"
MODALITY_COLUMNS <- c(AVERAGE_COLUMN, "Auditory", "Visual", "Motor")

MEASURE_COLOURS <- c(
  Reliability = "#0072B2",      # Okabe-Ito blue
  Confusability = "#D55E00",    # Okabe-Ito vermillion
  Distinctiveness = "#009E73"   # Okabe-Ito green
)

# Outlined-fill shapes: the fill and the outline are drawn as two layers so the
# pale fill (alpha 0.15) and the darker outline (alpha 0.6) can share one
# colour scale with the ribbons and model lines. The average column gets the
# neutral diamond, so it reads as a summary rather than as a fourth modality,
# and it is named in the legend rather than left unexplained.
MODALITY_SHAPES <- c(
  Average = 23,    # diamond
  Auditory = 21,   # circle
  Visual = 24,     # triangle
  Motor = 22       # square
)

# Every panel is named by its own strips, so neither figure carries a legend at
# all. The shape mapping is kept - it is what keeps the three modality columns
# visually distinct - but its guide is dropped: a shape legend would list the
# average alongside the modalities and imply it is a fourth one.
scale_measure_colour <- scale_colour_manual(
  values = MEASURE_COLOURS,
  guide = "none"
)
scale_measure_fill <- scale_fill_manual(
  values = MEASURE_COLOURS,
  guide = "none"
)
scale_modality_shape <- scale_shape_manual(
  values = MODALITY_SHAPES,
  breaks = MODALITY_COLUMNS,
  guide = "none"
)

MEASURE_STRIP_LABELS <- c(
  Reliability = "Neural reliability",
  Confusability = "Neural confusability",
  Distinctiveness = "Neural distinctiveness"
)
MODALITY_STRIP_LABELS <- c(
  Average = "All modalities",
  Auditory = "Auditory",
  Visual = "Visual",
  Motor = "Motor"
)
manuscript_labeller <- labeller(
  measure = MEASURE_STRIP_LABELS,
  modality = MODALITY_STRIP_LABELS
)

# Raw two-wave segments are drawn in neutral grey: they are the data layer, and
# leaving them uncoloured keeps colour reserved for the outcome and keeps the
# model line visible on top of them.
SEGMENT_COLOUR <- "grey55"
SEGMENT_ALPHA <- 0.25
SEGMENT_WIDTH <- 0.3

POINT_SIZE <- 1.2
POINT_STROKE <- 0.35
POINT_FILL_ALPHA <- 0.15
POINT_OUTLINE_ALPHA <- 0.6

RIBBON_ALPHA <- 0.25
LINE_WIDTH <- 0.8
GAP_LINE_WIDTH <- 0.6

# Ages and follow-up intervals are recorded in whole years, so points would
# otherwise stack into vertical stripes. Jitter is horizontal only - the
# outcome is never displaced - and is seeded so every figure and every rebuild
# places the same point in the same place.
JITTER_SEED <- 1
FIGURE2_JITTER_WIDTH <- 0.5
FIGURE3_JITTER_WIDTH <- 0.15

AGE_AXIS_LABEL <- "Age (years)"
WITHIN_TIME_LABEL <- "Years from first wave"
WITHIN_TIME_BREAKS <- c(0, 2, 4, 6)

FIGURE_WIDTH_MM <- 180
FIGURE_HEIGHT_MM <- 150       # one 3 x 4 grid, no stacked panels
FIGURE_DPI <- 300
GRID_POINTS <- 100

# The average column is set apart from the block of three modality columns by
# roughly one extra panel gap of whitespace. ggplot2 accepts one width per gap
# between panel columns, so this is the gap after column 1 followed by the gaps
# inside the modality block; widen or narrow the break by editing one number.
AVERAGE_COLUMN_GAP <- 16      # points (~5.6 mm at the saved figure width)
MODALITY_COLUMN_GAP <- 4      # points

# A thin rule is drawn in that gap, so the break survives greyscale print
# without tinting any panel: all four panels keep the same white background.
AVERAGE_DIVIDER_COLOUR <- "grey70"
AVERAGE_DIVIDER_WIDTH <- 0.5

manuscript_theme <- function(base_size = 9) {
  theme_classic(base_size = base_size) +
    theme(
      panel.grid = element_blank(),
      axis.text = element_text(colour = "black"),
      axis.title = element_text(colour = "black"),
      axis.line = element_line(colour = "black", linewidth = 0.3),
      axis.ticks = element_line(colour = "black", linewidth = 0.3),
      # Plain strip text at the axis-title size: the strips are labels, not
      # headings.
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text = element_text(face = "plain", size = rel(1), colour = "black"),
      legend.position = "none",
      legend.key.height = unit(9, "pt"),
      legend.margin = margin(t = 1, b = 0),
      plot.title = element_blank(),
      plot.subtitle = element_blank(),
      plot.tag = element_text(face = "bold", size = 11),
      plot.margin = margin(3, 5, 3, 3)
    )
}

# The average column's strip label is emphasised by weight alone: same
# typeface, same size, bold. Per-strip text needs ggh4x's themed strips, since
# theme(strip.text.x = ...) styles every column strip at once.
manuscript_strips <- function() {
  ggh4x::strip_themed(
    text_x = ggh4x::elem_list_text(
      face = c("bold", rep("plain", length(MODALITY_COLUMNS) - 1))
    )
  )
}

# The divider is a gtable edit, because no ggplot layer can draw in the gap
# between two panel columns - that space belongs to the layout, not to any
# panel. The rule is inserted into the existing spacing column, so no width
# changes and the panels keep the sizes ggplot2 gave them.
#
# Positions are looked up from the panel grobs rather than hard-coded: the gap
# is the widest layout column between panel column 1 and panel column 2, and
# the rule spans the panel rows only, so it stops short of the column strips
# above and the x axis below.
add_average_divider <- function(figure) {
  table <- ggplotGrob(figure)
  panels <- table$layout[grep("^panel", table$layout$name), ]

  panel_columns <- sort(unique(panels$l))
  panel_rows <- sort(unique(panels$t))
  stopifnot(length(panel_columns) == length(MODALITY_COLUMNS))

  gap_candidates <- seq(panel_columns[[1]] + 1, panel_columns[[2]] - 1)
  gap_widths <- vapply(
    gap_candidates,
    function(i) as.numeric(convertWidth(table$widths[i], "pt")),
    numeric(1)
  )
  gap_column <- gap_candidates[[which.max(gap_widths)]]
  # The gap must be the widened one, not a zero-width layout artefact.
  stopifnot(max(gap_widths) > MODALITY_COLUMN_GAP)

  widths_before <- table$widths
  table <- gtable::gtable_add_grob(
    table,
    linesGrob(
      x = unit(c(0.5, 0.5), "npc"),
      y = unit(c(0, 1), "npc"),
      gp = gpar(col = AVERAGE_DIVIDER_COLOUR, lwd = AVERAGE_DIVIDER_WIDTH)
    ),
    t = min(panel_rows),
    b = max(panel_rows),
    l = gap_column,
    r = gap_column,
    name = "average-divider"
  )
  # Inserting into an existing column cannot resize anything; assert it anyway,
  # because a shifted panel width would break the across-column comparison.
  stopifnot(identical(as.character(widths_before), as.character(table$widths)))

  table
}

# predict.merMod gives no standard errors, so the fixed-effect prediction and
# its Wald CI are computed from the design matrix and the fixed-effect vcov.
# `contrasts_option` must match the contrasts the model was fitted under.
fixed_effect_predictions <- function(model, newdata, contrasts_option = NULL) {
  if (!is.null(contrasts_option)) {
    previous_contrasts <- options(contrasts = contrasts_option)
    on.exit(options(previous_contrasts), add = TRUE)
  }

  design <- model.matrix(
    delete.response(terms(model, fixed.only = TRUE)),
    data = newdata
  )
  design <- design[, names(fixef(model)), drop = FALSE]

  fitted_values <- as.vector(design %*% fixef(model))
  standard_errors <- sqrt(
    rowSums((design %*% as.matrix(vcov(model))) * design)
  )
  critical_value <- qnorm(0.975)

  newdata %>%
    mutate(
      predicted = fitted_values,
      se = standard_errors,
      conf_low = fitted_values - critical_value * standard_errors,
      conf_high = fitted_values + critical_value * standard_errors
    )
}

complete_outcome_data <- function(data, outcome) {
  data %>% drop_na(all_of(c(outcome, MODEL_TERMS, "SubNum")))
}

# Both figures are drawn from one long frame with the same six columns, so the
# row order, the column order, and the column names are set in exactly one
# place and the two figures cannot drift apart.
as_plot_frame <- function(frame) {
  frame %>%
    transmute(
      participant,
      session,
      x,
      measure = factor(measure, levels = MEASURES),
      modality = factor(modality, levels = MODALITY_COLUMNS),
      value
    )
}

# Model lines carry the same measure/modality keys plus the grouping the
# solid/dashed split needs. line_group keeps each panel's ribbon and dashed
# span in one piece; solid_group splits the solid line at the unsampled gap.
as_line_frame <- function(frame) {
  frame %>%
    mutate(
      measure = factor(measure, levels = MEASURES),
      modality = factor(modality, levels = MODALITY_COLUMNS),
      line_group = paste(measure, modality),
      solid_group = paste(measure, modality, solid_segment)
    )
}

typical_covariates <- function(data) {
  list(
    Sex_M1 = median(data$Sex_M1, na.rm = TRUE),
    Education = median(data$Education, na.rm = TRUE)
  )
}

# No one was scanned between the young and older cohorts, so the age axis has
# an unsampled interval. The model line is drawn solid where there are data and
# dashed across the gap; the ribbon stays continuous. Ranges come from the data
# of that panel. `gap_span` includes both boundary points so the dashed span
# meets the solid spans exactly.
mark_age_gap <- function(grid_x, observed_ages) {
  young_ages <- observed_ages[observed_ages < OLDER_ADULT_AGE]
  older_ages <- observed_ages[observed_ages >= OLDER_ADULT_AGE]

  if (length(young_ages) == 0 || length(older_ages) == 0) {
    return(tibble(
      x = grid_x,
      solid_segment = "all",
      gap_span = FALSE
    ))
  }

  young_max <- max(young_ages, na.rm = TRUE)
  older_min <- min(older_ages, na.rm = TRUE)

  tibble(
    x = grid_x,
    solid_segment = case_when(
      grid_x <= young_max ~ "young",
      grid_x >= older_min ~ "older",
      TRUE ~ NA_character_
    ),
    gap_span = grid_x >= young_max & grid_x <= older_min
  )
}

# Figure 3 has no unsampled interval, so every point of the line is solid.
mark_no_gap <- function(grid_x) {
  tibble(x = grid_x, solid_segment = "all", gap_span = FALSE)
}

# One jitter offset per participant, applied to every observation that
# participant contributes, so both endpoints of a segment move together and the
# segment still meets its own points.
jitter_participant_x <- function(data, width, seed = JITTER_SEED) {
  participant_ids <- sort(unique(data$participant))
  offsets <- with_seeded_rng(seed, runif(length(participant_ids), -width, width))
  names(offsets) <- as.character(participant_ids)

  data %>%
    mutate(x = x + unname(offsets[as.character(participant)]))
}

# set.seed() inside a helper would leak into the caller's RNG state; this
# restores whatever stream was in use.
with_seeded_rng <- function(seed, expression) {
  if (exists(".Random.seed", envir = globalenv())) {
    previous_seed <- get(".Random.seed", envir = globalenv())
    on.exit(assign(".Random.seed", previous_seed, envir = globalenv()), add = TRUE)
  }
  set.seed(seed)
  expression
}

# ---------------------------------------------------------------------------
# Layer builders. Order is fixed everywhere: raw segments, then points, then
# the CI ribbon, then the model line on top of the data cloud.
# ---------------------------------------------------------------------------
segment_layer <- function(segments) {
  if (is.null(segments)) return(NULL)

  geom_line(
    data = segments,
    aes(x, value, group = participant),
    colour = SEGMENT_COLOUR, alpha = SEGMENT_ALPHA, linewidth = SEGMENT_WIDTH
  )
}

point_layers <- function(points, position) {
  list(
    geom_point(
      data = points,
      aes(x, value, fill = measure, shape = modality),
      colour = NA, alpha = POINT_FILL_ALPHA, size = POINT_SIZE,
      stroke = POINT_STROKE, position = position, show.legend = FALSE
    ),
    geom_point(
      data = points,
      aes(x, value, colour = measure, shape = modality),
      fill = NA, alpha = POINT_OUTLINE_ALPHA, size = POINT_SIZE,
      stroke = POINT_STROKE, position = position
    )
  )
}

model_layers <- function(lines) {
  list(
    geom_ribbon(
      data = lines,
      aes(x, ymin = conf_low, ymax = conf_high, fill = measure,
          group = line_group),
      alpha = RIBBON_ALPHA, colour = NA
    ),
    geom_line(
      data = lines %>% filter(gap_span),
      aes(x, predicted, colour = measure, group = line_group),
      linewidth = GAP_LINE_WIDTH, linetype = "dashed"
    ),
    geom_line(
      data = lines %>% filter(!is.na(solid_segment)),
      aes(x, predicted, colour = measure, group = solid_group),
      linewidth = LINE_WIDTH
    )
  )
}

# ---------------------------------------------------------------------------
# The figure builder. Both manuscript figures are this one 3 x 4 grid: measure
# is the row throughout, and the modality average is the leftmost column of the
# same grid rather than a separate panel stacked above it.
#
# scales = "free_y" is load-bearing. It frees the y scale by ROW and shares it
# within the row, so all four columns of a row sit on one scale, drawn once at
# the left; switch = "y" puts the measure strip where that y title would be.
# scales = "free" would give each of the twelve panels its own y axis and
# destroy the across-column comparison the row exists to support.
# ---------------------------------------------------------------------------
build_measure_modality_figure <- function(points, lines, x_label,
                                          segments = NULL,
                                          position = position_identity(),
                                          x_scale = NULL) {
  figure <- ggplot() +
    segment_layer(segments) +
    point_layers(points, position) +
    model_layers(lines) +
    # facet_grid2 is facet_grid plus per-strip theming; the faceting itself is
    # unchanged, including the load-bearing scales = "free_y".
    ggh4x::facet_grid2(
      measure ~ modality,
      scales = "free_y", switch = "y", labeller = manuscript_labeller,
      strip = manuscript_strips()
    ) +
    scale_measure_colour +
    scale_measure_fill +
    scale_modality_shape +
    guides(shape = "none") +
    labs(x = x_label, y = NULL) +
    manuscript_theme() +
    theme(
      panel.spacing.x = unit(
        c(
          AVERAGE_COLUMN_GAP,
          rep(MODALITY_COLUMN_GAP, length(MODALITY_COLUMNS) - 2)
        ),
        "pt"
      )
    )

  if (!is.null(x_scale)) figure <- figure + x_scale
  figure
}

# ---------------------------------------------------------------------------
# Figure 2: between-person age effects.
#
# The Average column is the former panel A, unchanged: the same participant
# means across modalities and waves, and the same pooled main-effect model line
# with its 95% CI ribbon. The three modality columns are the former panel B,
# unchanged: session x modality observations with the slopes implied by the
# interaction model.
# ---------------------------------------------------------------------------
figure2_average_points <- function(data) {
  map_dfr(MEASURES, function(measure_name) {
    complete_outcome_data(data, measure_name) %>%
      group_by(SubNum) %>%
      summarise(
        x = mean(Age_DuringParticipation_, na.rm = TRUE),
        value = mean(.data[[measure_name]], na.rm = TRUE),
        .groups = "drop"
      ) %>%
      transmute(
        participant = SubNum,
        # These points pool over waves as well as modalities, so they belong to
        # a participant rather than to any one session.
        session = NA_real_,
        x,
        measure = measure_name,
        modality = "Average",
        value
      )
  })
}

figure2_average_lines <- function(data, models) {
  map_dfr(MEASURES, function(measure_name) {
    column_data <- complete_outcome_data(data, measure_name)
    covariates <- typical_covariates(column_data)

    age_grid <- seq(
      min(column_data$Age_DuringParticipation_, na.rm = TRUE),
      max(column_data$Age_DuringParticipation_, na.rm = TRUE),
      length.out = GRID_POINTS
    )
    prediction_grid <- mark_age_gap(
      age_grid, column_data$Age_DuringParticipation_
    ) %>%
      mutate(
        between_cAge = x - OLDER_ADULT_AGE,
        within_cAge = 0,
        Sex_M1 = covariates$Sex_M1,
        Education = covariates$Education,
        # Effect-coded contrasts at zero: the average of the three modalities.
        aud_mot = 0,
        aud_vis = 0
      )

    fixed_effect_predictions(models[[measure_name]], prediction_grid) %>%
      mutate(measure = measure_name, modality = "Average")
  })
}

figure2_modality_points <- function(data) {
  map_dfr(MEASURES, function(measure_name) {
    prepare_interaction_data(complete_outcome_data(data, measure_name)) %>%
      transmute(
        participant = SubNum,
        session = Wave_Num,
        x = Age_DuringParticipation_,
        measure = measure_name,
        modality = as.character(Modality),
        value = .data[[measure_name]]
      )
  })
}

figure2_modality_lines <- function(data, interaction_models) {
  map_dfr(MEASURES, function(measure_name) {
    column_data <- prepare_interaction_data(
      complete_outcome_data(data, measure_name)
    )
    covariates <- typical_covariates(column_data)

    prediction_grid <- column_data %>%
      group_by(Modality) %>%
      group_modify(~ mark_age_gap(
        seq(
          min(.x$Age_DuringParticipation_, na.rm = TRUE),
          max(.x$Age_DuringParticipation_, na.rm = TRUE),
          length.out = GRID_POINTS
        ),
        .x$Age_DuringParticipation_
      )) %>%
      ungroup() %>%
      mutate(
        between_cAge = x - OLDER_ADULT_AGE,
        within_cAge = 0,
        Sex_M1 = covariates$Sex_M1,
        Education = covariates$Education
      )

    # Modality stays a factor with the interaction model's own levels until the
    # design matrix has been built; only then does it become a plotting column.
    fixed_effect_predictions(
      interaction_models[[measure_name]],
      prediction_grid,
      contrasts_option = INTERACTION_CONTRASTS
    ) %>%
      mutate(measure = measure_name, modality = as.character(Modality))
  })
}

figure2_plot_data <- function(data, models, interaction_models) {
  list(
    points = bind_rows(
      figure2_average_points(data),
      figure2_modality_points(data)
    ) %>%
      as_plot_frame(),
    lines = bind_rows(
      figure2_average_lines(data, models),
      figure2_modality_lines(data, interaction_models)
    ) %>%
      as_line_frame()
  )
}

# Takes the frame built above rather than building it, so the figure and the
# checks run on exactly the same rows.
build_figure2 <- function(figure_data) {
  build_measure_modality_figure(
    figure_data$points,
    figure_data$lines,
    AGE_AXIS_LABEL,
    position = position_jitter(
      width = FIGURE2_JITTER_WIDTH, height = 0, seed = JITTER_SEED
    )
  )
}

# ---------------------------------------------------------------------------
# Figure 3: within-person age effects, on time since each participant's own
# first retained wave so every trajectory starts at zero.
#
# within_cAge is centred on each participant's mean age, so it maps onto the
# plotted axis with a constant offset. The line is clipped to the follow-up
# actually observed in that panel; being a straight line, its slope and CI are
# unaffected by where it is truncated.
# ---------------------------------------------------------------------------
within_prediction_grid <- function(data) {
  offset <- mean(data$x - data$within_cAge, na.rm = TRUE)
  within_range <- range(data$within_cAge, na.rm = TRUE) + offset
  time_range <- c(
    max(0, within_range[1]),
    min(max(data$x, na.rm = TRUE), within_range[2])
  )

  grid_x <- seq(time_range[1], time_range[2], length.out = GRID_POINTS)
  mark_no_gap(grid_x) %>%
    mutate(within_cAge = x - offset)
}

# The session frames keep `within_cAge` alongside the six plotting columns
# because the prediction grids are built from it; as_plot_frame() drops it
# again before anything is drawn.
figure3_average_sessions <- function(data) {
  map_dfr(MEASURES, function(measure_name) {
    complete_outcome_data(data, measure_name) %>%
      group_by(SubNum, Wave_Num) %>%
      summarise(
        age = mean(Age_DuringParticipation_, na.rm = TRUE),
        within_cAge = mean(within_cAge, na.rm = TRUE),
        value = mean(.data[[measure_name]], na.rm = TRUE),
        .groups = "drop"
      ) %>%
      group_by(SubNum) %>%
      filter(n() >= 2) %>%
      mutate(x = age - min(age, na.rm = TRUE)) %>%
      ungroup() %>%
      transmute(
        participant = SubNum,
        session = Wave_Num,
        x,
        within_cAge,
        measure = measure_name,
        modality = "Average",
        value
      )
  })
}

figure3_average_lines <- function(data, models, sessions) {
  map_dfr(MEASURES, function(measure_name) {
    covariates <- typical_covariates(complete_outcome_data(data, measure_name))

    prediction_grid <- sessions %>%
      filter(measure == measure_name) %>%
      within_prediction_grid() %>%
      mutate(
        between_cAge = 0,
        Sex_M1 = covariates$Sex_M1,
        Education = covariates$Education,
        aud_mot = 0,
        aud_vis = 0
      )

    fixed_effect_predictions(models[[measure_name]], prediction_grid) %>%
      mutate(measure = measure_name, modality = "Average")
  })
}

figure3_modality_sessions <- function(data) {
  map_dfr(MEASURES, function(measure_name) {
    prepare_interaction_data(complete_outcome_data(data, measure_name)) %>%
      group_by(SubNum, Modality) %>%
      filter(n_distinct(Wave_Num) >= 2) %>%
      mutate(
        x = Age_DuringParticipation_ -
          min(Age_DuringParticipation_, na.rm = TRUE)
      ) %>%
      ungroup() %>%
      transmute(
        participant = SubNum,
        session = Wave_Num,
        x,
        within_cAge,
        measure = measure_name,
        modality = as.character(Modality),
        value = .data[[measure_name]]
      )
  })
}

figure3_modality_lines <- function(data, interaction_models, sessions) {
  map_dfr(MEASURES, function(measure_name) {
    covariates <- typical_covariates(complete_outcome_data(data, measure_name))

    prediction_grid <- sessions %>%
      filter(measure == measure_name) %>%
      group_by(modality) %>%
      group_modify(~ within_prediction_grid(.x)) %>%
      ungroup() %>%
      mutate(
        between_cAge = 0,
        Sex_M1 = covariates$Sex_M1,
        Education = covariates$Education,
        Modality = factor(modality, levels = INTERACTION_MODALITY_LEVELS)
      )

    fixed_effect_predictions(
      interaction_models[[measure_name]],
      prediction_grid,
      contrasts_option = INTERACTION_CONTRASTS
    ) %>%
      mutate(measure = measure_name)
  })
}

figure3_plot_data <- function(data, models, interaction_models) {
  # The prediction grids are built from the unjittered sessions; the display
  # frame is jittered once - not per layer - so each participant's segment and
  # its endpoint markers all move by the same offset.
  average_sessions <- figure3_average_sessions(data)
  modality_sessions <- figure3_modality_sessions(data)

  list(
    points = bind_rows(average_sessions, modality_sessions) %>%
      as_plot_frame() %>%
      jitter_participant_x(FIGURE3_JITTER_WIDTH),
    lines = bind_rows(
      figure3_average_lines(data, models, average_sessions),
      figure3_modality_lines(data, interaction_models, modality_sessions)
    ) %>%
      as_line_frame()
  )
}

build_figure3 <- function(figure_data) {
  # Points and grey connecting segments are the same rows, so they cannot come
  # apart: the segment joins exactly the two markers it belongs to.
  build_measure_modality_figure(
    figure_data$points,
    figure_data$lines,
    WITHIN_TIME_LABEL,
    segments = figure_data$points,
    x_scale = scale_x_continuous(breaks = WITHIN_TIME_BREAKS)
  )
}

# ---------------------------------------------------------------------------
# Checks on the Average column.
#
# The Average column is a summary of the modality columns, so it is recomputed
# from the session x modality observations and compared rather than trusted.
# Motion exclusion is modality-dependent, so a session can survive with fewer
# than three modalities; those sessions are averages over an incomplete set and
# are counted here.
# ---------------------------------------------------------------------------
modality_observations <- function(data) {
  map_dfr(MEASURES, function(measure_name) {
    prepare_interaction_data(complete_outcome_data(data, measure_name)) %>%
      transmute(
        participant = SubNum,
        session = Wave_Num,
        modality = as.character(Modality),
        measure = measure_name,
        value = .data[[measure_name]]
      )
  })
}

session_modality_means <- function(observations) {
  observations %>%
    group_by(measure, participant, session) %>%
    summarise(
      session_value = mean(value),
      n_modalities = n_distinct(modality),
      .groups = "drop"
    )
}

average_column_values <- function(points) {
  points %>%
    filter(modality == "Average") %>%
    mutate(measure = as.character(measure)) %>%
    select(measure, participant, session, plotted = value)
}

# Figure 3's Average points are session-level, so they are exactly the
# per-session mean across the modalities retained for that session.
figure3_average_check <- function(data, points) {
  sessions <- session_modality_means(modality_observations(data))
  plotted <- average_column_values(points)

  comparison <- plotted %>%
    inner_join(sessions, by = c("measure", "participant", "session"))
  stopifnot(nrow(comparison) == nrow(plotted))

  plotted_sessions <- sessions %>%
    semi_join(plotted, by = c("measure", "participant", "session"))

  list(
    figure = "Figure 3",
    average_definition = "per-session mean across retained modalities",
    rows_compared = nrow(comparison),
    max_difference = max(abs(comparison$plotted - comparison$session_value)),
    session_cells = nrow(plotted_sessions),
    incomplete_cells = sum(plotted_sessions$n_modalities < 3),
    incomplete_sessions = plotted_sessions %>%
      filter(n_modalities < 3) %>%
      distinct(participant, session) %>%
      nrow()
  )
}

# Figure 2's Average points pool over waves as well, so the comparison is made
# at the participant level; the extra term reports how far that participant
# mean sits from the unweighted mean of its own session means, which differ
# only when a participant's sessions retained different numbers of modalities.
figure2_average_check <- function(data, points) {
  observations <- modality_observations(data)
  sessions <- session_modality_means(observations)
  plotted <- average_column_values(points)

  participant_means <- observations %>%
    group_by(measure, participant) %>%
    summarise(observation_mean = mean(value), .groups = "drop")
  session_based_means <- sessions %>%
    group_by(measure, participant) %>%
    summarise(session_mean = mean(session_value), .groups = "drop")

  comparison <- plotted %>%
    inner_join(participant_means, by = c("measure", "participant")) %>%
    inner_join(session_based_means, by = c("measure", "participant"))
  stopifnot(nrow(comparison) == nrow(plotted))

  list(
    figure = "Figure 2",
    average_definition = "participant mean across retained sessions x modalities",
    rows_compared = nrow(comparison),
    max_difference = max(abs(comparison$plotted - comparison$observation_mean)),
    max_session_based_difference =
      max(abs(comparison$plotted - comparison$session_mean)),
    session_cells = nrow(sessions),
    incomplete_cells = sum(sessions$n_modalities < 3),
    incomplete_sessions = sessions %>%
      filter(n_modalities < 3) %>%
      distinct(participant, session) %>%
      nrow()
  )
}

print_average_check <- function(check) {
  cat(check$figure, "- Average column\n")
  cat("  definition:", check$average_definition, "\n")
  cat(sprintf(
    "  %d Average points recomputed from the modality columns; max |difference| = %.3g\n",
    check$rows_compared, check$max_difference
  ))
  if (!is.null(check$max_session_based_difference)) {
    cat(sprintf(
      "  max |plotted - unweighted mean of that participant's session means| = %.3g\n",
      check$max_session_based_difference
    ))
  }
  cat(sprintf(
    "  session x measure cells: %d; averaged over fewer than 3 modalities: %d (%.1f%%)\n",
    check$session_cells, check$incomplete_cells,
    100 * check$incomplete_cells / check$session_cells
  ))
  cat(sprintf(
    "  distinct sessions averaged over fewer than 3 modalities: %d\n",
    check$incomplete_sessions
  ))
}

# ---------------------------------------------------------------------------
# Checks on the grid itself, run before anything is written to disk: the grid
# must be 3 x 4 with no empty panel, and the four panels of a row must share
# one y scale (which is what scales = "free_y" buys and scales = "free" would
# throw away).
# ---------------------------------------------------------------------------
verify_manuscript_figure <- function(figure, points) {
  panel_counts <- points %>%
    count(measure, modality, .drop = FALSE)

  stopifnot(
    nrow(panel_counts) == length(MEASURES) * length(MODALITY_COLUMNS),
    all(panel_counts$n > 0)
  )

  built <- ggplot_build(figure)
  layout <- built$layout$layout
  stopifnot(nrow(layout) == length(MEASURES) * length(MODALITY_COLUMNS))

  y_limits <- map_dfr(seq_len(nrow(layout)), function(i) {
    range <- built$layout$panel_params[[i]]$y.range
    tibble(
      row = layout$ROW[[i]],
      measure = as.character(layout$measure[[i]]),
      modality = as.character(layout$modality[[i]]),
      y_min = range[[1]],
      y_max = range[[2]]
    )
  })

  row_spread <- y_limits %>%
    group_by(row) %>%
    summarise(
      spread = max(abs(y_min - first(y_min))) + max(abs(y_max - first(y_max))),
      .groups = "drop"
    )
  stopifnot(all(row_spread$spread < 1e-9))

  list(panels = panel_counts, y_limits = y_limits)
}

save_manuscript_figure <- function(figure, directory, name,
                                   height_mm = FIGURE_HEIGHT_MM) {
  dir.create(directory, showWarnings = FALSE, recursive = TRUE)

  ggsave(
    file.path(directory, paste0(name, ".pdf")),
    figure,
    width = FIGURE_WIDTH_MM, height = height_mm, units = "mm"
  )
  ggsave(
    file.path(directory, paste0(name, ".png")),
    figure,
    width = FIGURE_WIDTH_MM, height = height_mm, units = "mm", dpi = FIGURE_DPI
  )

  file.path(directory, paste0(name, c(".pdf", ".png")))
}
```

```{r manuscript-figures, include=FALSE}
figure_dir <- file.path(output_dir, "figures")
supplement_figure_dir <- file.path(figure_dir, "supplement")

manuscript_figures <- lapply(MOTION_RULES, function(rule) {
  rule_data <- analysis_data_by_rule[[rule$key]]
  rule_models <- fit_outcome_models(rule_data)
  rule_interaction_models <- interaction_results_by_rule[[rule$key]]$models

  figure2_data <- figure2_plot_data(
    rule_data, rule_models, rule_interaction_models
  )
  figure3_data <- figure3_plot_data(
    rule_data, rule_models, rule_interaction_models
  )

  figure2_plot <- build_figure2(figure2_data)
  figure3_plot <- build_figure3(figure3_data)

  # The checks run on the ggplot, before it is turned into a gtable: the layout
  # checks need ggplot_build(), which a gtable no longer supports. Nothing is
  # written until they pass; they stop() on failure.
  figure2_layout <- verify_manuscript_figure(figure2_plot, figure2_data$points)
  figure3_layout <- verify_manuscript_figure(figure3_plot, figure3_data$points)

  # The saved and displayed figures are the gtables, divider included.
  list(
    figure2 = add_average_divider(figure2_plot),
    figure3 = add_average_divider(figure3_plot),
    figure2_layout = figure2_layout,
    figure3_layout = figure3_layout,
    figure2_average = figure2_average_check(rule_data, figure2_data$points),
    figure3_average = figure3_average_check(rule_data, figure3_data$points)
  )
})
names(manuscript_figures) <- vapply(
  MOTION_RULES, `[[`, character(1), "key"
)

saved_figure_files <- c(
  save_manuscript_figure(
    manuscript_figures$primary$figure2, figure_dir, "figure2"
  ),
  save_manuscript_figure(
    manuscript_figures$primary$figure3, figure_dir, "figure3"
  ),
  save_manuscript_figure(
    manuscript_figures$liberal$figure2, supplement_figure_dir, "figure2"
  ),
  save_manuscript_figure(
    manuscript_figures$liberal$figure3, supplement_figure_dir, "figure3"
  )
)
```

## Figure 2 (primary motion rule)

```{r figure2-display, fig.width=7.09, fig.height=5.91, out.width='100%'}
# The figures are gtables once the divider has been added, and printing a
# gtable writes its layout to the console instead of drawing it.
grid.newpage()
grid.draw(manuscript_figures$primary$figure2)
```

## Figure 3 (primary motion rule)

```{r figure3-display, fig.width=7.09, fig.height=5.91, out.width='100%'}
grid.newpage()
grid.draw(manuscript_figures$primary$figure3)
```

## Average-column checks (primary motion rule)

The Average column is recomputed from the modality columns and compared with
what is plotted. Motion exclusion is modality-dependent, so the count of
sessions averaged over fewer than three modalities is reported alongside.

```{r average-column-check}
renderPrint({
  print_average_check(manuscript_figures$primary$figure2_average)
  cat("\n")
  print_average_check(manuscript_figures$primary$figure3_average)
})
```

```{r figure-files}
renderTable({
  tibble(
    File = sub(
      paste0("^", project_dir, .Platform$file.sep), "", saved_figure_files
    )
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# Interactive age-effect figure

The combined figure contains auditory, motor, and visual ROI panels. The blue
and red lines are the model-implied cross-sectional and longitudinal age
effects from ROI-specific mixed-effects models adjusted for sex and education.
Each panel reports the corresponding unstandardized coefficient (*B*),
standardized coefficient (β), and *p* value.

```{r plot-helper, include=FALSE}
format_age_p <- function(p) {
  if (is.na(p)) "NA" else if (p < 0.001) "< .001" else sprintf("%.3f", p)
}

get_roi_age_plot_info <- function(data, modality, outcome) {
  plot_data <- data %>%
    filter(Modality == modality) %>%
    select(
      Subject, SubNum, Wave_Num, Age_DuringParticipation_, n_waves,
      mean_age, within_cAge, between_cAge, Sex_M1, Education,
      Modality, all_of(outcome)
    ) %>%
    rename(value = all_of(outcome)) %>%
    drop_na(
      value, Age_DuringParticipation_, within_cAge, between_cAge,
      Sex_M1, Education, SubNum
    )

  roi_model <- lmer(
    reformulate(
      c(
        "within_cAge", "between_cAge", "Sex_M1", "Education",
        "(1 | SubNum)"
      ),
      response = "value"
    ),
    data = plot_data,
    REML = FALSE
  )

  coefficients <- summary(roi_model)$coefficients
  outcome_sd <- sd(plot_data$value, na.rm = TRUE)
  within_b <- coefficients["within_cAge", "Estimate"]
  between_b <- coefficients["between_cAge", "Estimate"]
  within_beta <- within_b * sd(plot_data$within_cAge, na.rm = TRUE) / outcome_sd
  between_beta <- between_b * sd(plot_data$between_cAge, na.rm = TRUE) / outcome_sd
  stats_label <- sprintf(
    paste0(
      "Long.: B=%.3f, β=%.3f, p=%s\n",
      "Cross-sec.: B=%.3f, β=%.3f, p=%s"
    ),
    within_b,
    within_beta,
    format_age_p(coefficients["within_cAge", "Pr(>|t|)"]),
    between_b,
    between_beta,
    format_age_p(coefficients["between_cAge", "Pr(>|t|)"])
  )

  longitudinal_data <- plot_data %>% filter(n_waves > 1)
  typical_sex <- median(plot_data$Sex_M1, na.rm = TRUE)
  typical_education <- median(plot_data$Education, na.rm = TRUE)

  between_age <- seq(
    min(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    max(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    length.out = 100
  )
  between_line <- tibble(
    age = between_age,
    within_cAge = 0,
    between_cAge = between_age - OLDER_ADULT_AGE,
    Sex_M1 = typical_sex,
    Education = typical_education,
    SubNum = plot_data$SubNum[1]
  ) %>%
    mutate(predicted = predict(
      roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
    ))

  if (nrow(longitudinal_data) > 0 &&
      diff(range(longitudinal_data$within_cAge, na.rm = TRUE)) > 0) {
    within_values <- seq(
      min(longitudinal_data$within_cAge, na.rm = TRUE),
      max(longitudinal_data$within_cAge, na.rm = TRUE),
      length.out = 100
    )
    within_age <- mean(longitudinal_data$mean_age, na.rm = TRUE) +
      within_values
    within_line <- tibble(
      age = within_age,
      within_cAge = within_values,
      between_cAge = 0,
      Sex_M1 = typical_sex,
      Education = typical_education,
      SubNum = plot_data$SubNum[1]
    ) %>%
      mutate(predicted = predict(
        roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
      ))
  } else {
    within_line <- tibble(age = numeric(), predicted = numeric())
  }

  list(
    data = plot_data,
    longitudinal_data = longitudinal_data,
    between_line = between_line,
    within_line = within_line,
    stats_label = stats_label
  )
}

make_roi_age_panel <- function(data, modality, outcome, label) {
  plot_info <- get_roi_age_plot_info(data, modality, outcome)
  roi_labels <- c(Aud = "Auditory", Mot = "Motor", Vis = "Visual")

  point_data <- plot_info$data

  figure <- ggplot(point_data, aes(Age_DuringParticipation_, value)) +
    geom_line(
      data = plot_info$longitudinal_data %>% arrange(SubNum, Wave_Num),
      aes(group = SubNum),
      colour = "#CC0000", linewidth = 0.35, alpha = 0.45
    ) +
    geom_point(size = 1.25, alpha = 0.5) +
    geom_line(
      data = plot_info$between_line,
      aes(age, predicted, colour = "Cross-sectional"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = plot_info$within_line,
      aes(age, predicted, colour = "Longitudinal"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    scale_colour_manual(
      values = c("Cross-sectional" = "#3366CC", "Longitudinal" = "#CC0000"),
      name = NULL
    ) +
    labs(
      title = unname(roi_labels[[modality]]),
      subtitle = plot_info$stats_label,
      x = "Age",
      y = paste("Neural", tolower(outcome)),
      caption = label
    ) +
    theme_classic(base_size = 10) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 6.8, hjust = 0.5, lineheight = 1.1),
      plot.caption = element_text(size = 8.5, hjust = 0.5),
      axis.title = element_text(size = 9.5),
      axis.text = element_text(size = 8.5),
      legend.text = element_text(size = 8.5)
    )

  list(
    figure = figure,
    roi_label = unname(roi_labels[[modality]]),
    stats_label = plot_info$stats_label
  )
}
```

```{r age-figure, fig.width=15, fig.height=5.2, out.width='100%'}
renderPlot({
  req(input$outcome)

  modalities <- c("Aud", "Mot", "Vis")
  panels <- lapply(
    modalities,
    function(modality) {
      make_roi_age_panel(
        analysis_data(),
        modality,
        input$outcome,
        ""
      )
    }
  )

  wrap_plots(lapply(panels, `[[`, "figure"), nrow = 1) +
    plot_layout(guides = "collect") +
    plot_annotation(
      title = paste(input$outcome, "age effects"),
      subtitle = rule_label(),
      theme = theme(
        plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 10)
      )
    ) &
    theme(legend.position = "bottom")
}, res = 110)
```

# Excluded runs

```{r excluded-runs}
renderTable({
  selected_flags() %>%
    filter(exclude) %>%
    arrange(desc(pct_excluded), Subject, Modality) %>%
    select(Subject, Modality, pct_excluded) %>%
    rename(`Percent excluded TRs` = pct_excluded)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```
