---
title: "Motion-Corrected Analysis (Interactive)"
author: "Tien"
date: "2026-07-26"
output:
  html_document:
    toc: true
    toc_float: true
runtime: shiny
---

```{r setup, include=FALSE}
# The pipeline lives in R/ and is shared with the numbered scripts in script/,
# so this viewer and the files in output/ cannot drift apart. Nothing in this
# document writes to disk; script/01..06 do that.
source(here::here("R", "setup.R"))
source(here("R", "data.R"))
source(here("R", "models.R"))
source(here("R", "sensitivity.R"))
source(here("R", "magnitude.R"))
source(here("R", "figures.R"))

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

<style>
.main-container {
  max-width: 1500px !important;
}
.shiny-plot-output {
  width: 100% !important;
}
</style>

# Choose the motion-exclusion rule

Choose the framewise-displacement (FD) cutoff and the percentage of excluded
TRs that causes a run to be removed. The models, sample summary, and plots
below update automatically. The exported CSVs are always written for both
manuscript rules (FD 0.4 mm / 10% and FD 0.5 mm / 20%) regardless of the
selection here.

```{r controls}
fluidRow(
  column(
    width = 5,
    selectInput(
      "fd_cutoff",
      "FD cutoff",
      choices = c("0.5 mm" = "0.5", "0.4 mm" = "0.4"),
      selected = "0.4"
    )
  ),
  column(
    width = 7,
    sliderInput(
      "pct_cutoff",
      "Exclude a run when percentage of excluded TRs is at least:",
      min = 0,
      max = 100,
      value = 10,
      step = 1,
      post = "%"
    )
  )
)

helpText(
  "The primary rule is 0.4 mm / 10%; the liberal rule is 0.5 mm / 20%."
)

observeEvent(input$fd_cutoff, {
  updateSliderInput(
    session,
    "pct_cutoff",
    value = if (identical(input$fd_cutoff, "0.4")) 10 else 20
  )
}, ignoreInit = TRUE)
```

# Data preparation

Only Waves 1 and 2 are analysed. Wave 3 (MIND-C) rows are dropped from the
prepared data before anything else happens, so the motion-exclusion accounting,
the demographics, and the models all describe the same Waves 1--2 sample.

```{r build-data, include=FALSE}
# Built straight from data/raw/, so the document does not depend on
# script/01_prepare_data.R having been run first.
prepared <- build_all_from_source()
analysis_data_by_rule <- prepared$analysis_data_by_rule
run_flags_by_rule <- prepared$run_flags_by_rule
```

```{r reactive-analysis, include=FALSE}
selected_flags <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  build_run_flags(input$fd_cutoff, input$pct_cutoff)
})

analysis_data <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  build_analysis_data(input$fd_cutoff, input$pct_cutoff)
})

missing_motion_records <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  count_missing_motion_records(input$fd_cutoff, input$pct_cutoff)
})

models <- reactive({
  fit_outcome_models(analysis_data())
})

rule_label <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  rule_text(input$fd_cutoff, input$pct_cutoff)
})
```

# Sample after exclusion

```{r sample-summary}
renderTable({
  flags <- selected_flags()
  kept <- analysis_data()

  tibble(
    `Selected rule` = rule_label(),
    `Runs assessed` = nrow(flags),
    `Runs flagged` = sum(flags$exclude, na.rm = TRUE),
    `Runs flagged (%)` = sprintf(
      "%.1f%%",
      100 * mean(flags$exclude, na.rm = TRUE)
    ),
    `Unique scans retained` = n_distinct(kept$Subject),
    `Unique participants retained` = n_distinct(kept$SubNum),
    `Modality-level observations retained` = nrow(kept),
    `Rows dropped for a missing motion record` = missing_motion_records()
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Missing motion records

`filter(!exclude)` drops any session x modality that has no run-level motion
record at all, because such a row was never assessed for motion. This is the
intended behaviour: it reproduces the sample sizes reported in the manuscript
(724 observations under the primary rule, 841 under the liberal rule). The
earlier `filter(is.na(exclude) | !exclude)` retained unassessed rows and gives
larger samples (808 and 925 respectively). The count below is the number of
healthy Waves 1--2 rows removed for a *missing* record rather than for an
actual motion exclusion; it does not depend on the FD or percentage cutoff and
should be described in the manuscript.

```{r missing-motion-records}
renderTable({
  bind_rows(lapply(MOTION_RULES, function(rule) {
    joined <- build_joined_data(rule$fd_cutoff, rule$pct_cutoff) %>%
      filter(Subgroup == 1)
    tibble(
      `Motion rule` = rule$label,
      `Healthy Waves 1-2 rows before motion filtering` = nrow(joined),
      `Rows with no motion record` = sum(is.na(joined$exclude)),
      `Rows excluded for motion` = sum(joined$exclude, na.rm = TRUE),
      `Rows retained` = sum(!is.na(joined$exclude) & !joined$exclude)
    )
  }))
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# APA-style demographics and descriptive statistics

These tables use the currently selected motion-exclusion rule. The
cross-sectional sample is defined as Wave 1. The longitudinal sample requires
retained observations at both Waves 1 and 2; Wave 1 demographics are reported
so each longitudinal participant is counted once. Young adults are below age
65 and older adults are age 65 or above. In Table 2, longitudinal descriptive
statistics average each participant's two retained waves within modality
before calculating the group mean and standard deviation.

```{r apa-table-data, include=FALSE}
crosssec_data <- reactive({
  analysis_data() %>%
    filter(Wave_Num == 1) %>%
    mutate(
      Age_Group = factor(
        if_else(
          Age_DuringParticipation_ < OLDER_ADULT_AGE,
          "Young adults",
          "Older adults"
        ),
        levels = c("Young adults", "Older adults")
      )
    )
})

demographics_summary <- reactive({
  crosssec_participants <- crosssec_data() %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Cross-sectional")

  longitudinal_ids <- analysis_data() %>%
    distinct(SubNum, Wave_Num) %>%
    count(SubNum, name = "n_retained_waves") %>%
    filter(n_retained_waves >= 2) %>%
    pull(SubNum)

  longitudinal_participants <- crosssec_data() %>%
    filter(SubNum %in% longitudinal_ids) %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_participants, longitudinal_participants) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group) %>%
    summarise(
      n = n(),
      Age_M = mean(Age_DuringParticipation_, na.rm = TRUE),
      Age_SD = sd(Age_DuringParticipation_, na.rm = TRUE),
      Education_M = mean(Education, na.rm = TRUE),
      Education_SD = sd(Education, na.rm = TRUE),
      Male_n = sum(Sex_M1 == 1, na.rm = TRUE),
      Female_n = sum(Sex_M1 == 2, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      `Age, M (SD)` = sprintf("%.2f (%.2f)", Age_M, Age_SD),
      `Education, M (SD)` = sprintf(
        "%.2f (%.2f)",
        Education_M,
        Education_SD
      ),
      `Male, n (%)` = sprintf("%d (%.1f%%)", Male_n, 100 * Male_n / n),
      `Female, n (%)` = sprintf(
        "%d (%.1f%%)",
        Female_n,
        100 * Female_n / n
      )
    ) %>%
    select(
      Sample,
      `Age group` = Age_Group,
      N = n,
      `Age, M (SD)`,
      `Education, M (SD)`,
      `Male, n (%)`,
      `Female, n (%)`
    )
})

descriptives_summary <- reactive({
  crosssec_descriptives <- crosssec_data() %>%
    group_by(SubNum, Age_Group, Modality) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(Sample = "Cross-sectional")

  baseline_age_groups <- crosssec_data() %>%
    distinct(SubNum, Age_Group)

  longitudinal_descriptives <- analysis_data() %>%
    group_by(SubNum, Modality) %>%
    filter(n_distinct(Wave_Num) >= 2) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    inner_join(baseline_age_groups, by = "SubNum") %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_descriptives, longitudinal_descriptives) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group, Modality) %>%
    summarise(
      Reliability_M = mean(Reliability, na.rm = TRUE),
      Reliability_SD = sd(Reliability, na.rm = TRUE),
      Confusability_M = mean(Confusability, na.rm = TRUE),
      Confusability_SD = sd(Confusability, na.rm = TRUE),
      Distinctiveness_M = mean(Distinctiveness, na.rm = TRUE),
      Distinctiveness_SD = sd(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      Modality = recode(
        as.character(Modality),
        Aud = "Auditory",
        Mot = "Motor",
        Vis = "Visual"
      )
    ) %>%
    arrange(
      Sample,
      Age_Group,
      factor(Modality, levels = c("Auditory", "Motor", "Visual"))
    )
})

style_apa_table <- function(ft) {
  ft %>%
    theme_booktabs() %>%
    font(fontname = "Times New Roman", part = "all") %>%
    fontsize(size = 11, part = "all") %>%
    bold(part = "header") %>%
    align(align = "center", part = "all") %>%
    align(j = 1, align = "left", part = "all") %>%
    padding(padding = 4, part = "all") %>%
    autofit()
}

demographics_flextable <- reactive({
  flextable(demographics_summary()) %>%
    merge_v(j = "Sample") %>%
    valign(j = "Sample", valign = "top") %>%
    style_apa_table()
})

descriptives_flextable <- reactive({
  flextable(descriptives_summary()) %>%
    colformat_double(
      j = c(
        "Reliability_M", "Reliability_SD",
        "Confusability_M", "Confusability_SD",
        "Distinctiveness_M", "Distinctiveness_SD"
      ),
      digits = 3,
      na_str = "—"
    ) %>%
    set_header_labels(
      Sample = "Sample",
      Age_Group = "Age group",
      Modality = "Modality",
      Reliability_M = "M",
      Reliability_SD = "SD",
      Confusability_M = "M",
      Confusability_SD = "SD",
      Distinctiveness_M = "M",
      Distinctiveness_SD = "SD"
    ) %>%
    add_header_row(
      values = c(
        "Sample", "Age group", "Modality", "Reliability",
        "Confusability", "Distinctiveness"
      ),
      colwidths = c(1, 1, 1, 2, 2, 2)
    ) %>%
    merge_v(j = "Sample") %>%
    merge_v(
      j = c("Sample", "Age_Group"),
      target = "Age_Group",
      combine = TRUE
    ) %>%
    valign(
      j = c("Sample", "Age_Group"),
      valign = "top"
    ) %>%
    style_apa_table()
})

apa_table_html <- function(number, title, note, ft) {
  htmltools::tagList(
    htmltools::tags$p(
      htmltools::tags$strong(number),
      style = "font-family: 'Times New Roman'; margin-bottom: 0;"
    ),
    htmltools::tags$p(
      htmltools::tags$em(title),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 1.05em; margin-top: 0;"
      )
    ),
    flextable::htmltools_value(ft),
    htmltools::tags$p(
      htmltools::tags$em("Note. "),
      paste0(note, " Selected rule: ", rule_label(), "."),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 0.95em; margin-top: 8px;"
      )
    )
  )
}
```

## Table 1

```{r demographics-table}
renderUI({
  apa_table_html(
    "Table 1",
    "Cross-Sectional and Longitudinal Analysis Demographics Summary",
    paste(
      "Values are presented as M (SD) unless otherwise indicated.",
      "The cross-sectional sample includes retained Wave 1 participants.",
      "The longitudinal sample includes participants with retained",
      "observations at Waves 1 and 2; their Wave 1 demographics are shown.",
      "Participants may therefore appear in both samples."
    ),
    demographics_flextable()
  )
})

downloadButton("download_demographics", "Download Table 1 (.docx)")

output$download_demographics <- downloadHandler(
  filename = function() "table1_demographics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 1. Cross-Sectional and Longitudinal Demographics" =
        demographics_flextable(),
      path = file
    )
  }
)
```

## Table 2

```{r descriptives-table}
renderUI({
  apa_table_html(
    "Table 2",
    "Cross-Sectional and Longitudinal Analysis Descriptive Statistics",
    paste(
      "Values are means and standard deviations for reliability",
      "(within-category similarity), confusability (between-category",
      "similarity), and distinctiveness by sample, age group, and modality.",
      "Longitudinal values are participant-level averages across two",
      "retained waves within each modality."
    ),
    descriptives_flextable()
  )
})

downloadButton("download_descriptives", "Download Table 2 (.docx)")

output$download_descriptives <- downloadHandler(
  filename = function() "table2_descriptive_statistics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 2. Cross-Sectional and Longitudinal Descriptives" =
        descriptives_flextable(),
      path = file
    )
  }
)
```

# Mixed-effects models

The fixed-effects specification is unchanged from the original document:
within-person age, between-person age, sex, education, and the two
effect-coded modality contrasts, with a participant random intercept.

```{r model-choice}
selectInput(
  "outcome",
  "Outcome shown below",
  choices = c("Reliability", "Confusability", "Distinctiveness"),
  selected = "Reliability"
)
```

## Model summary

```{r model-summary}
renderPrint({
  cat(rule_label(), "\n\n")
  print(summary(models()[[input$outcome]]))
})
```

## Type III tests

Fixed effects are tested with Satterthwaite-approximated denominator degrees of
freedom, consistent with the interaction tests below.

```{r model-anova}
renderPrint({
  cat(rule_label(), "\n\n")
  print(stats::anova(
    models()[[input$outcome]],
    type = 3,
    ddf = "Satterthwaite"
  ))
})
```

# Modality-specific age effects

Each outcome is refit with the two age terms crossed with modality:

```
outcome ~ (between_cAge + within_cAge) * Modality +
          Sex_M1 + Education + (1 | SubNum)
```

`Modality` enters as a factor (Auditory, Visual, Motor) rather than the manual
`aud_mot` / `aud_vis` contrasts, and `contr.sum` is in effect for these fits so
the Type III tests are valid. The joint interaction tests are the 2-df
Satterthwaite *F* tests for `between_cAge:Modality` and `within_cAge:Modality`
from `anova()`.

Simple slopes are probed with `emmeans::emtrends()` on the interaction model —
not with stratified per-modality models — and pairwise slope differences use
Holm adjustment across the three modality pairs. All degrees of freedom are
Satterthwaite.

Everything is computed for both age terms and all three outcomes so the exports
are complete. The `interaction_significant` column marks the joint *F* test at
*p* < .05, which is the gate the manuscript's pre-registered rule uses: only a
significant interaction is decomposed and interpreted.

```{r interaction-models, include=FALSE}
interaction_results_by_rule <- lapply(
  analysis_data_by_rule,
  compute_interaction_results
)

interaction_results <- reactive({
  compute_interaction_results(analysis_data())
})
```

## Joint interaction tests

```{r interaction-tests-table}
renderTable({
  format_interaction_tests(interaction_results()$tests)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Simple slopes by modality

```{r interaction-slopes-table}
renderTable({
  format_simple_slopes(interaction_results()$slopes)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Pairwise slope differences (Holm)

```{r interaction-contrasts-table}
renderTable({
  format_slope_contrasts(interaction_results()$contrasts)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# Magnitude of the age effects

The magnitude contrast, $\Delta = |b_{\mathrm{reliability}}| -
|b_{\mathrm{confusability}}|$, asks which component is more strongly associated
with age on their shared correlation scale; positive values indicate a larger
absolute reliability slope. Unlike the signed component difference, it does
not become large merely because the slopes point in opposite directions.
Motor cortex illustrates this: opposite signs and comparable between-person
magnitudes accompany a steep distinctiveness slope but a magnitude contrast
near zero.

The point estimates below follow the FD and percentage controls because only
four model fits are needed. The bootstrap intervals and p values are precomputed
for the two manuscript rules only and do not respond to the controls; refitting
thousands of participant resamples would be too slow for interactive use.
The signed difference is computed from the two component models; it need not
exactly equal the coefficient from a separate distinctiveness mixed model,
which estimates its own variance components.

```{r magnitude-data, include=FALSE}
selected_magnitude <- reactive({
  component_magnitude_estimates(analysis_data())
})

# Read precomputed inference once; never run or save a bootstrap in the viewer.
magnitude_path <- file.path(TABLE_DIR, "magnitude_comparison.csv")
precomputed_magnitude <- if (file.exists(magnitude_path)) {
  readr::read_csv(magnitude_path, num_threads = 1, show_col_types = FALSE)
} else NULL
```

## Point estimates for the selected rule

```{r magnitude-point-table}
renderTable({
  selected_magnitude() %>%
    mutate(across(c(b_reliability, b_confusability, signed_diff, delta),
                  ~ sprintf("%.5f", .x)))
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Bootstrap comparison for both manuscript rules

These are the saved manuscript-rule results, independent of the controls above.
`n_boot_ok` gives the number of usable draws; a bootstrap p value of zero means
that no usable resample crossed zero, rather than an exactly zero probability.

```{r magnitude-bootstrap-table}
renderUI({
  if (is.null(precomputed_magnitude)) {
    return(helpText(paste(
      "Bootstrap results are not available yet.",
      "Run script/04_magnitude_comparison.R to compute them."
    )))
  }
  htmltools::tagList(lapply(MOTION_RULES, function(rule) {
    table <- precomputed_magnitude %>%
      filter(motion_rule == rule$label) %>%
      transmute(
        Scope = scope, `Age term` = age_term,
        `Reliability b` = sprintf("%.5f", b_reliability),
        `Confusability b` = sprintf("%.5f", b_confusability),
        `Signed difference` = sprintf("%.5f", signed_diff),
        Delta = sprintf("%.5f", delta),
        `95% percentile CI` = sprintf("[%.5f, %.5f]", delta_ci_lo, delta_ci_hi),
        `Bootstrap p` = format_p(boot_p), n_boot_ok
      )
    htmltools::tagList(
      htmltools::tags$h4(rule$label),
      htmltools::HTML(knitr::kable(
        table, format = "html", row.names = FALSE,
        table.attr = "class='table table-striped table-bordered'"
      ))
    )
  }))
})
```

# Contextual effect: do the within- and between-person age slopes differ?

The main models estimate a within-person slope (`within_cAge`) and a
between-person slope (`between_cAge`) separately, but neither *t* test asks
whether the two differ from each other. That difference, `b_between -
b_within`, is the contextual effect of Hoffman and Stawski (2009); a
contextual effect indistinguishable from zero means one common age slope
describes both the cross-sectional and the longitudinal change.

It is computed here two ways, and the two must agree:

1. **Reparameterisation.** `age_session_c = within_cAge + between_cAge` is
   session age centred at `r OLDER_ADULT_AGE`, so replacing the pair
   (`within_cAge`, `between_cAge`) with (`age_session_c`, `between_cAge`)
   spans the same column space and is the same model refit in new
   coordinates. In those coordinates the slope on `age_session_c` *is* the
   within-person slope and the coefficient on `between_cAge` *is* the
   contextual effect, so its Satterthwaite *t* test is the test of
   H0: `b_between = b_within`.
2. **Contrast on the original model.** `lmerTest::contest1D()` with
   *L* = -1 on `within_cAge`, +1 on `between_cAge`, zero elsewhere.

Both routes use the estimator settings of the main models (`REML = FALSE`,
Satterthwaite df) and the same complete-case rows. The chunk below asserts the
agreement rather than trusting it: the reparameterised fit must reproduce the
original within-person slope and log-likelihood, and the two routes must return
the same estimate, SE, df, *t*, and *p*.

```{r contextual-effects, include=FALSE}
contextual_effects_by_rule <- setNames(
  lapply(MOTION_RULES, function(rule) {
    contextual_effects_table(analysis_data_by_rule[[rule$key]], rule$label)
  }),
  rule_keys()
)
```

## Contextual effects for both manuscript rules

These do not depend on the controls above; they are the numbers written to
`contextual_effects.csv`.

```{r contextual-effects-table}
renderUI({
  htmltools::tagList(lapply(MOTION_RULES, function(rule) {
    htmltools::tagList(
      htmltools::tags$h4(rule$label),
      htmltools::HTML(
        knitr::kable(
          format_contextual_effects(contextual_effects_by_rule[[rule$key]]),
          format = "html",
          row.names = FALSE,
          table.attr = "class='table table-striped table-bordered'"
        )
      )
    )
  }))
})
```

# Outlier sensitivity (MAD x 3)

Outliers are flagged with a median absolute deviation criterion applied to each
outcome **separately within each modality**, because the auditory, motor, and
visual distributions differ enough that pooling them inflates the MAD and
flags almost nothing. An observation is flagged when
`abs(x - median(x)) > 3 * mad(x)`; R's `mad()` already applies the 1.4826
consistency constant.

Two exclusion variants are then compared with the full sample. A row is treated
as flagged if it is an outlier on *any* of the three outcomes, so all three
models are refit on the same subsample and the estimates stay comparable
across outcomes.

- **Drop flagged observations** removes only the flagged session x modality rows.
- **Drop flagged participants** removes every row belonging to a participant
  with at least one flagged observation.

The model specification is unchanged: `outcome ~ within_cAge + between_cAge +
Sex_M1 + Education + aud_mot + aud_vis + (1 | SubNum)`, `REML = FALSE`, with
Satterthwaite *p* values from `lmerTest`.

```{r outlier-sensitivity, include=FALSE}
outlier_sensitivity_by_rule <- setNames(
  lapply(MOTION_RULES, function(rule) {
    outlier_sensitivity_table(analysis_data_by_rule[[rule$key]], rule$label)
  }),
  rule_keys()
)
```

## Flagged observations under the selected rule

```{r outlier-counts}
renderPrint({
  data <- analysis_data()
  counts <- mad_outlier_counts(data)
  cat(rule_label(), "\n")
  cat("Analytic observations:", nrow(data), "\n")
  cat("Participants:", n_distinct(data$SubNum), "\n\n")
  for (i in seq_len(nrow(counts))) {
    cat(sprintf(
      "%-16s %3d observations flagged (%.1f%%) across %d participants\n",
      counts$Outcome[[i]],
      counts$`Observations flagged`[[i]],
      counts$`Observations flagged (%)`[[i]],
      counts$`Participants affected`[[i]]
    ))
  }
})
```

## Comparison table under the selected rule

```{r outlier-sensitivity-table}
renderTable({
  format_outlier_sensitivity(outlier_sensitivity_table(analysis_data()))
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Comparison tables for both manuscript rules

These two tables do not depend on the controls above; they are the numbers
written to `outlier_sensitivity.csv`.

```{r outlier-sensitivity-both-rules}
renderUI({
  htmltools::tagList(lapply(MOTION_RULES, function(rule) {
    table <- outlier_sensitivity_by_rule[[rule$key]]
    htmltools::tagList(
      htmltools::tags$h4(rule$label),
      htmltools::HTML(
        knitr::kable(
          format_outlier_sensitivity(table),
          format = "html",
          row.names = FALSE,
          table.attr = "class='table table-striped table-bordered'"
        )
      )
    )
  }))
})
```

# Exported numbers for the manuscript

This document writes nothing. The tables the manuscript needs are written to
`output/tables/` by `script/02_fit_models.R`, `script/03_sensitivity.R`, and
`script/04_magnitude_comparison.R`, and the figures to `output/figures/` by
`script/05_figures.R`. Those scripts and this viewer call the same functions in `R/`, so the numbers shown below
are the numbers in those files.

# Manuscript figures

Figures 2 and 3 are static, publication-ready files. `script/05_figures.R`
writes both under both motion rules to `output/figures/`, with the rule in the
filename: `figure2_fd04_pct10.png` is the primary rule and
`figure2_fd05_pct20.png` the liberal one, which is the supplementary figure.
Shown below is the primary rule. This section does not depend on the Shiny
controls.

Both figures are one 3 x 4 grid built by the same function, so they cannot
drift apart. Measure is the row throughout — reliability, confusability,
distinctiveness — and the four columns are the modality average, labelled
*All modalities*, followed by auditory, visual, and motor. The average is the
leftmost column of the same grid rather than a separate panel stacked above it,
and it is marked as the primary result three ways: an extra panel gap of
whitespace after it, a bold strip label, and a thin grey rule drawn in that
gap, which survives greyscale printing. All four panels keep the same white
background, and nothing else about the column differs — point size, alpha, line
width, ribbon opacity, and colour are identical across all four columns, so the
panels in a row stay directly comparable.

The grid uses `scales = "free_y"`, which frees the y scale by row and shares it
within the row: all four columns of a row sit on one scale, drawn once at the
left, so the average and the three modalities are directly comparable.
`scales = "free"` would give each of the twelve panels its own axis and destroy
that comparison, so it is not used. `switch = "y"` puts the measure strip where
the y title would otherwise be.

In Figure 2 the x-axis is age. The average column shows the pooled
between-person age effect from the main (no-interaction) model with a 95% CI
ribbon, over participant-level values averaged across modalities and waves, so
each participant appears once — session x modality points are unreadable once
pooled. The three modality columns show the same effect resolved by modality
using the slopes implied by the interaction model, so the lines correspond
exactly to the emmeans simple slopes reported above; there the points are the
individual session x modality observations. No one was scanned between the
young and older cohorts, so every model line is solid where there are data and
dashed across the unsampled 30--60 region, with the ribbon continuous
throughout.

Figure 3 has the same 3 x 4 structure for the within-person effect, on an
x-axis of years from each participant's own first retained wave rather than
chronological age, so every trajectory starts at zero. Only returning
participants appear. The average column is the pooled within-person slope from
the main model with a 95% CI ribbon, over session-level values averaged across
modalities; the modality columns resolve it by modality using the interaction
model, matching the emmeans `within_cAge` simple slopes. Their ribbons overlap
heavily across modalities, which is the honest display of a within-person x
modality interaction that is non-significant for all three outcomes. The grey
lines join each participant's own two sessions.

Both figures share one aesthetic mapping, defined once and reused by every
panel. Colour encodes the measure and nothing else (Okabe-Ito blue, vermillion,
and green for reliability, confusability, and distinctiveness) and is applied
to points, model lines, and CI ribbons. Shape encodes the column and nothing
else — open circle, triangle, and square for auditory, visual, and motor, and
a neutral open diamond for the average. Neither figure carries a legend: every
panel is named by its own strips, and a shape legend would list the average
alongside the three modalities and imply it is a fourth one. The grey two-wave
segments in Figure 3 are the raw data layer and are deliberately left
uncoloured so the model line stays visible on top of them.

Titles, subtitles, and in-panel statistics are deliberately absent: panel
identity comes from the strip labels, and the statistics belong in Table 3 and
the manuscript text.

```{r manuscript-figures, include=FALSE}
# Built for display only; script/05_figures.R writes the files.
manuscript_figures <- lapply(MOTION_RULES, function(rule) {
  rule_data <- analysis_data_by_rule[[rule$key]]
  rule_models <- fit_outcome_models(rule_data)
  rule_interaction_models <- interaction_results_by_rule[[rule$key]]$models

  figure2_data <- figure2_plot_data(
    rule_data, rule_models, rule_interaction_models
  )
  figure3_data <- figure3_plot_data(
    rule_data, rule_models, rule_interaction_models
  )

  figure2_plot <- build_figure2(figure2_data)
  figure3_plot <- build_figure3(figure3_data)

  # Same checks the script runs; they stop() on failure.
  verify_manuscript_figure(figure2_plot, figure2_data$points)
  verify_manuscript_figure(figure3_plot, figure3_data$points)

  list(
    figure2 = add_average_divider(figure2_plot),
    figure3 = add_average_divider(figure3_plot),
    figure2_average = figure2_average_check(rule_data, figure2_data$points),
    figure3_average = figure3_average_check(rule_data, figure3_data$points)
  )
})
names(manuscript_figures) <- rule_keys()
```

## Figure 2 (primary motion rule)

```{r figure2-display, fig.width=7.09, fig.height=5.91, out.width='100%'}
# The figures are gtables once the divider has been added, and printing a
# gtable writes its layout to the console instead of drawing it.
grid.newpage()
grid.draw(manuscript_figures$primary$figure2)
```

## Figure 3 (primary motion rule)

```{r figure3-display, fig.width=7.09, fig.height=5.91, out.width='100%'}
grid.newpage()
grid.draw(manuscript_figures$primary$figure3)
```

## Average-column checks (primary motion rule)

The Average column is recomputed from the modality columns and compared with
what is plotted. Motion exclusion is modality-dependent, so the count of
sessions averaged over fewer than three modalities is reported alongside.

```{r average-column-check}
renderPrint({
  print_average_check(manuscript_figures$primary$figure2_average)
  cat("\n")
  print_average_check(manuscript_figures$primary$figure3_average)
})
```

# Interactive age-effect figure

The combined figure contains auditory, motor, and visual ROI panels. The blue
and red lines are the model-implied cross-sectional and longitudinal age
effects from ROI-specific mixed-effects models adjusted for sex and education.
Each panel reports the corresponding unstandardized coefficient (*B*),
standardized coefficient (β), and *p* value.

```{r plot-helper, include=FALSE}
format_age_p <- function(p) {
  if (is.na(p)) "NA" else if (p < 0.001) "< .001" else sprintf("%.3f", p)
}

get_roi_age_plot_info <- function(data, modality, outcome) {
  plot_data <- data %>%
    filter(Modality == modality) %>%
    select(
      Subject, SubNum, Wave_Num, Age_DuringParticipation_, n_waves,
      mean_age, within_cAge, between_cAge, Sex_M1, Education,
      Modality, all_of(outcome)
    ) %>%
    rename(value = all_of(outcome)) %>%
    drop_na(
      value, Age_DuringParticipation_, within_cAge, between_cAge,
      Sex_M1, Education, SubNum
    )

  roi_model <- lmer(
    reformulate(
      c(
        "within_cAge", "between_cAge", "Sex_M1", "Education",
        "(1 | SubNum)"
      ),
      response = "value"
    ),
    data = plot_data,
    REML = FALSE
  )

  coefficients <- summary(roi_model)$coefficients
  outcome_sd <- sd(plot_data$value, na.rm = TRUE)
  within_b <- coefficients["within_cAge", "Estimate"]
  between_b <- coefficients["between_cAge", "Estimate"]
  within_beta <- within_b * sd(plot_data$within_cAge, na.rm = TRUE) / outcome_sd
  between_beta <- between_b * sd(plot_data$between_cAge, na.rm = TRUE) / outcome_sd
  stats_label <- sprintf(
    paste0(
      "Long.: B=%.3f, β=%.3f, p=%s\n",
      "Cross-sec.: B=%.3f, β=%.3f, p=%s"
    ),
    within_b,
    within_beta,
    format_age_p(coefficients["within_cAge", "Pr(>|t|)"]),
    between_b,
    between_beta,
    format_age_p(coefficients["between_cAge", "Pr(>|t|)"])
  )

  longitudinal_data <- plot_data %>% filter(n_waves > 1)
  typical_sex <- median(plot_data$Sex_M1, na.rm = TRUE)
  typical_education <- median(plot_data$Education, na.rm = TRUE)

  between_age <- seq(
    min(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    max(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    length.out = 100
  )
  between_line <- tibble(
    age = between_age,
    within_cAge = 0,
    between_cAge = between_age - OLDER_ADULT_AGE,
    Sex_M1 = typical_sex,
    Education = typical_education,
    SubNum = plot_data$SubNum[1]
  ) %>%
    mutate(predicted = predict(
      roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
    ))

  if (nrow(longitudinal_data) > 0 &&
      diff(range(longitudinal_data$within_cAge, na.rm = TRUE)) > 0) {
    within_values <- seq(
      min(longitudinal_data$within_cAge, na.rm = TRUE),
      max(longitudinal_data$within_cAge, na.rm = TRUE),
      length.out = 100
    )
    within_age <- mean(longitudinal_data$mean_age, na.rm = TRUE) +
      within_values
    within_line <- tibble(
      age = within_age,
      within_cAge = within_values,
      between_cAge = 0,
      Sex_M1 = typical_sex,
      Education = typical_education,
      SubNum = plot_data$SubNum[1]
    ) %>%
      mutate(predicted = predict(
        roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
      ))
  } else {
    within_line <- tibble(age = numeric(), predicted = numeric())
  }

  list(
    data = plot_data,
    longitudinal_data = longitudinal_data,
    between_line = between_line,
    within_line = within_line,
    stats_label = stats_label
  )
}

make_roi_age_panel <- function(data, modality, outcome, label) {
  plot_info <- get_roi_age_plot_info(data, modality, outcome)
  roi_labels <- c(Aud = "Auditory", Mot = "Motor", Vis = "Visual")

  point_data <- plot_info$data

  figure <- ggplot(point_data, aes(Age_DuringParticipation_, value)) +
    geom_line(
      data = plot_info$longitudinal_data %>% arrange(SubNum, Wave_Num),
      aes(group = SubNum),
      colour = "#CC0000", linewidth = 0.35, alpha = 0.45
    ) +
    geom_point(size = 1.25, alpha = 0.5) +
    geom_line(
      data = plot_info$between_line,
      aes(age, predicted, colour = "Cross-sectional"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = plot_info$within_line,
      aes(age, predicted, colour = "Longitudinal"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    scale_colour_manual(
      values = c("Cross-sectional" = "#3366CC", "Longitudinal" = "#CC0000"),
      name = NULL
    ) +
    labs(
      title = unname(roi_labels[[modality]]),
      subtitle = plot_info$stats_label,
      x = "Age",
      y = paste("Neural", tolower(outcome)),
      caption = label
    ) +
    theme_classic(base_size = 10) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 6.8, hjust = 0.5, lineheight = 1.1),
      plot.caption = element_text(size = 8.5, hjust = 0.5),
      axis.title = element_text(size = 9.5),
      axis.text = element_text(size = 8.5),
      legend.text = element_text(size = 8.5)
    )

  list(
    figure = figure,
    roi_label = unname(roi_labels[[modality]]),
    stats_label = plot_info$stats_label
  )
}
```

```{r age-figure, fig.width=15, fig.height=5.2, out.width='100%'}
renderPlot({
  req(input$outcome)

  modalities <- c("Aud", "Mot", "Vis")
  panels <- lapply(
    modalities,
    function(modality) {
      make_roi_age_panel(
        analysis_data(),
        modality,
        input$outcome,
        ""
      )
    }
  )

  wrap_plots(lapply(panels, `[[`, "figure"), nrow = 1) +
    plot_layout(guides = "collect") +
    plot_annotation(
      title = paste(input$outcome, "age effects"),
      subtitle = rule_label(),
      theme = theme(
        plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 10)
      )
    ) &
    theme(legend.position = "bottom")
}, res = 110)
```

# Excluded runs

```{r excluded-runs}
renderTable({
  selected_flags() %>%
    filter(exclude) %>%
    arrange(desc(pct_excluded), Subject, Modality) %>%
    select(Subject, Modality, pct_excluded) %>%
    rename(`Percent excluded TRs` = pct_excluded)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```
