---
title: "Motion-Corrected Analysis (Interactive)"
author: "Tien"
date: "2026-07-26"
output:
  html_document:
    toc: true
    toc_float: true
runtime: shiny
---

```{r setup, include=FALSE}
required_packages <- c(
  "shiny", "tidyverse", "lme4", "lmerTest", "car", "patchwork",
  "flextable"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install these packages before running the document: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(shiny)
library(tidyverse)
library(lme4)
library(lmerTest)
library(patchwork)
library(flextable)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

<style>
.main-container {
  max-width: 1500px !important;
}
.shiny-plot-output {
  width: 100% !important;
}
</style>

# Choose the motion-exclusion rule

Choose the framewise-displacement (FD) cutoff and the percentage of excluded
TRs that causes a run to be removed. The models, sample summary, and plots
below update automatically. The exported CSVs are always written for both
manuscript rules (FD 0.4 mm / 10% and FD 0.5 mm / 20%) regardless of the
selection here.

```{r controls}
fluidRow(
  column(
    width = 5,
    selectInput(
      "fd_cutoff",
      "FD cutoff",
      choices = c("0.5 mm" = "0.5", "0.4 mm" = "0.4"),
      selected = "0.4"
    )
  ),
  column(
    width = 7,
    sliderInput(
      "pct_cutoff",
      "Exclude a run when percentage of excluded TRs is at least:",
      min = 0,
      max = 100,
      value = 10,
      step = 1,
      post = "%"
    )
  )
)

helpText(
  "The primary rule is 0.4 mm / 10%; the liberal rule is 0.5 mm / 20%."
)

observeEvent(input$fd_cutoff, {
  updateSliderInput(
    session,
    "pct_cutoff",
    value = if (identical(input$fd_cutoff, "0.4")) 10 else 20
  )
}, ignoreInit = TRUE)
```

# Data preparation

Only Waves 1 and 2 are analysed. Wave 3 (MIND-C) rows are dropped from the
prepared data before anything else happens, so the motion-exclusion accounting,
the demographics, and the models all describe the same Waves 1--2 sample.

```{r locate-data, include=FALSE}
# The document reads the prepared outcome files and the run-level motion
# summaries. Set MIND_DATA_DIR (or MIND_PROJECT_DIR, whose "data" subdirectory
# is used) to override the search; otherwise the repository's data/ directory is
# found relative to the working directory, which is the script/ folder when the
# document is launched with runtime: shiny.
required_data_files <- c(
  "df_mind_all.csv",
  "df_mind_all_mc04.csv",
  "motion_exclusion_by_run_5mm.csv",
  "motion_exclusion_by_run_4mm.csv"
)

candidate_data_dirs <- c(
  Sys.getenv("MIND_DATA_DIR", unset = ""),
  file.path(Sys.getenv("MIND_PROJECT_DIR", unset = ""), "data"),
  file.path(Sys.getenv("MIND_PROJECT_DIR", unset = ""), "Data", "Motion"),
  "data",
  file.path("..", "data"),
  file.path("..", "..", "data")
)
candidate_data_dirs <- candidate_data_dirs[nzchar(candidate_data_dirs)]

has_required_files <- function(path) {
  dir.exists(path) &&
    all(file.exists(file.path(path, required_data_files)))
}

data_dir <- candidate_data_dirs[vapply(
  candidate_data_dirs, has_required_files, logical(1)
)]
if (length(data_dir) == 0) {
  stop(
    "Could not find the MiND data directory. Set MIND_DATA_DIR to the folder ",
    "containing ", paste(required_data_files, collapse = ", "), "."
  )
}
data_dir <- normalizePath(data_dir[[1]], mustWork = TRUE)
project_dir <- normalizePath(file.path(data_dir, ".."), mustWork = TRUE)
output_dir <- file.path(project_dir, "output")
```

```{r load-and-prepare, include=FALSE}
# Waves 1 and 2 only. The exploratory three-wave MIND-C analysis has been
# removed from the manuscript and its supplement, so Wave 3 never enters the
# pipeline.
ANALYSIS_WAVES <- c(1, 2)
OUTCOMES <- c("Reliability", "Confusability", "Distinctiveness")
MODEL_TERMS <- c(
  "within_cAge", "between_cAge", "Sex_M1", "Education",
  "aud_mot", "aud_vis"
)
OLDER_ADULT_AGE <- 65
MAD_THRESHOLD <- 3

# The prepared outcome files hold the 2,000-vertex motion-corrected measures:
# df_mind_all.csv is the FD 0.5 mm scrubbing pipeline and df_mind_all_mc04.csv
# the FD 0.4 mm pipeline. Each is keyed to the matching run-level motion file.
outcome_files <- c(
  "0.5" = "df_mind_all.csv",
  "0.4" = "df_mind_all_mc04.csv"
)
motion_files <- c(
  "0.5" = "motion_exclusion_by_run_5mm.csv",
  "0.4" = "motion_exclusion_by_run_4mm.csv"
)

read_data_file <- function(filename) {
  readr::read_csv(
    file.path(data_dir, filename),
    num_threads = 1,
    show_col_types = FALSE
  )
}

prepare_outcome_data <- function(data) {
  data %>%
    filter(!is.na(SubNum)) %>%
    pivot_longer(
      cols = matches("^(Aud|Mot|Vis)(Distinctiveness|Reliability|Confusability)$"),
      names_to = c("Modality", ".value"),
      names_pattern = "(Aud|Mot|Vis)(.*)"
    ) %>%
    mutate(Modality = factor(Modality, levels = c("Aud", "Mot", "Vis"))) %>%
    # Wave 3 is dropped here, i.e. before the motion-exclusion accounting, so
    # the run counts reported in the manuscript are Waves 1-2 only.
    filter(Wave_Num %in% ANALYSIS_WAVES)
}

outcome_data <- lapply(outcome_files, function(filename) {
  prepare_outcome_data(read_data_file(filename))
})

motion_exclusions <- lapply(motion_files, read_data_file)

healthy_subject_ids <- bind_rows(outcome_data) %>%
  filter(Subgroup == 1) %>%
  distinct(Subject) %>%
  pull(Subject)

subject_age_groups <- bind_rows(outcome_data) %>%
  filter(Subgroup == 1) %>%
  distinct(Subject, Age_DuringParticipation_) %>%
  transmute(
    Subject,
    Age_Group = factor(
      if_else(
        Age_DuringParticipation_ < OLDER_ADULT_AGE,
        "Young adults",
        "Older adults"
      ),
      levels = c("Young adults", "Older adults")
    )
  ) %>%
  distinct(Subject, Age_Group)
```

```{r analysis-pipeline, include=FALSE}
# Everything below is written as plain functions of (fd_cutoff, pct_cutoff) so
# the reactive display and the CSV export can share one implementation.

build_run_flags <- function(fd_cutoff, pct_cutoff) {
  motion_exclusions[[fd_cutoff]] %>%
    filter(
      subject %in% healthy_subject_ids,
      session == "placebo",
      run == "001",
      experiment %in% c("auditory", "visual", "motor")
    ) %>%
    transmute(
      Subject = subject,
      Modality = recode(
        experiment,
        auditory = "Aud",
        motor = "Mot",
        visual = "Vis"
      ),
      pct_excluded,
      exclude = pct_excluded >= pct_cutoff
    )
}

build_joined_data <- function(fd_cutoff, pct_cutoff) {
  outcome_data[[fd_cutoff]] %>%
    left_join(
      build_run_flags(fd_cutoff, pct_cutoff),
      by = c("Subject", "Modality")
    )
}

build_motion_filtered_data <- function(fd_cutoff, pct_cutoff) {
  build_joined_data(fd_cutoff, pct_cutoff) %>%
    # Intended behaviour, confirmed against the reported sample sizes: a
    # session x modality with NO run-level motion record is DROPPED, not
    # retained. `!exclude` already drops NA, but the NA test is written out so
    # the intent is explicit and is not mistaken for the earlier
    # `is.na(exclude) | !exclude` behaviour, which kept unassessed rows. See
    # "Missing motion records" below for how many rows this affects.
    filter(!is.na(exclude), !exclude)
}

count_missing_motion_records <- function(fd_cutoff, pct_cutoff) {
  build_joined_data(fd_cutoff, pct_cutoff) %>%
    filter(Subgroup == 1, is.na(exclude)) %>%
    nrow()
}

add_age_decomposition <- function(data) {
  data %>%
    group_by(SubNum) %>%
    mutate(
      n_waves = n_distinct(Wave_Num),
      mean_age = if_else(
        n_waves > 1,
        mean(Age_DuringParticipation_, na.rm = TRUE),
        Age_DuringParticipation_
      ),
      within_cAge = if_else(
        n_waves > 1,
        Age_DuringParticipation_ - mean_age,
        0
      ),
      between_cAge = mean_age - OLDER_ADULT_AGE,
      aud_mot = case_when(
        Modality == "Aud" ~ -1,
        Modality == "Mot" ~ 1,
        TRUE ~ 0
      ),
      aud_vis = case_when(
        Modality == "Aud" ~ -1,
        Modality == "Vis" ~ 1,
        TRUE ~ 0
      )
    ) %>%
    ungroup()
}

build_analysis_data <- function(fd_cutoff, pct_cutoff) {
  build_motion_filtered_data(fd_cutoff, pct_cutoff) %>%
    filter(Subgroup == 1, Wave_Num %in% ANALYSIS_WAVES) %>%
    add_age_decomposition()
}

fit_model <- function(data, outcome) {
  model_data <- data %>%
    drop_na(all_of(c(outcome, MODEL_TERMS, "SubNum")))

  if (nrow(model_data) == 0) {
    stop("No complete observations are available for ", outcome)
  }

  lmer(
    reformulate(
      c(MODEL_TERMS, "(1 | SubNum)"),
      response = outcome
    ),
    data = model_data,
    REML = FALSE
  )
}

fit_outcome_models <- function(data) {
  setNames(
    lapply(OUTCOMES, function(outcome) fit_model(data, outcome)),
    OUTCOMES
  )
}

rule_text <- function(fd_cutoff, pct_cutoff) {
  paste0(
    "FD ≥ ", fd_cutoff, " mm; run excluded at ≥ ",
    pct_cutoff, "% excluded TRs"
  )
}

# The two rules the manuscript reports. `prefix` is used for rule-specific
# export filenames.
MOTION_RULES <- list(
  list(
    key = "primary",
    fd_cutoff = "0.4",
    pct_cutoff = 10,
    label = "FD 0.4 mm / 10% (primary)",
    prefix = "fd04_pct10"
  ),
  list(
    key = "liberal",
    fd_cutoff = "0.5",
    pct_cutoff = 20,
    label = "FD 0.5 mm / 20% (liberal)",
    prefix = "fd05_pct20"
  )
)

format_p <- function(p) {
  if_else(
    is.na(p),
    NA_character_,
    if_else(p < .001, "< .001", sprintf("%.3f", p))
  )
}
```

```{r reactive-analysis, include=FALSE}
selected_flags <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  build_run_flags(input$fd_cutoff, input$pct_cutoff)
})

analysis_data <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  build_analysis_data(input$fd_cutoff, input$pct_cutoff)
})

missing_motion_records <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  count_missing_motion_records(input$fd_cutoff, input$pct_cutoff)
})

models <- reactive({
  fit_outcome_models(analysis_data())
})

rule_label <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)
  rule_text(input$fd_cutoff, input$pct_cutoff)
})
```

# Sample after exclusion

```{r sample-summary}
renderTable({
  flags <- selected_flags()
  kept <- analysis_data()

  tibble(
    `Selected rule` = rule_label(),
    `Runs assessed` = nrow(flags),
    `Runs flagged` = sum(flags$exclude, na.rm = TRUE),
    `Runs flagged (%)` = sprintf(
      "%.1f%%",
      100 * mean(flags$exclude, na.rm = TRUE)
    ),
    `Unique scans retained` = n_distinct(kept$Subject),
    `Unique participants retained` = n_distinct(kept$SubNum),
    `Modality-level observations retained` = nrow(kept),
    `Rows dropped for a missing motion record` = missing_motion_records()
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Missing motion records

`filter(!exclude)` drops any session x modality that has no run-level motion
record at all, because such a row was never assessed for motion. This is the
intended behaviour: it reproduces the sample sizes reported in the manuscript
(724 observations under the primary rule, 841 under the liberal rule). The
earlier `filter(is.na(exclude) | !exclude)` retained unassessed rows and gives
larger samples (808 and 925 respectively). The count below is the number of
healthy Waves 1--2 rows removed for a *missing* record rather than for an
actual motion exclusion; it does not depend on the FD or percentage cutoff and
should be described in the manuscript.

```{r missing-motion-records}
renderTable({
  bind_rows(lapply(MOTION_RULES, function(rule) {
    joined <- build_joined_data(rule$fd_cutoff, rule$pct_cutoff) %>%
      filter(Subgroup == 1)
    tibble(
      `Motion rule` = rule$label,
      `Healthy Waves 1-2 rows before motion filtering` = nrow(joined),
      `Rows with no motion record` = sum(is.na(joined$exclude)),
      `Rows excluded for motion` = sum(joined$exclude, na.rm = TRUE),
      `Rows retained` = sum(!is.na(joined$exclude) & !joined$exclude)
    )
  }))
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# APA-style demographics and descriptive statistics

These tables use the currently selected motion-exclusion rule. The
cross-sectional sample is defined as Wave 1. The longitudinal sample requires
retained observations at both Waves 1 and 2; Wave 1 demographics are reported
so each longitudinal participant is counted once. Young adults are below age
65 and older adults are age 65 or above. In Table 2, longitudinal descriptive
statistics average each participant's two retained waves within modality
before calculating the group mean and standard deviation.

```{r apa-table-data, include=FALSE}
crosssec_data <- reactive({
  analysis_data() %>%
    filter(Wave_Num == 1) %>%
    mutate(
      Age_Group = factor(
        if_else(
          Age_DuringParticipation_ < OLDER_ADULT_AGE,
          "Young adults",
          "Older adults"
        ),
        levels = c("Young adults", "Older adults")
      )
    )
})

demographics_summary <- reactive({
  crosssec_participants <- crosssec_data() %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Cross-sectional")

  longitudinal_ids <- analysis_data() %>%
    distinct(SubNum, Wave_Num) %>%
    count(SubNum, name = "n_retained_waves") %>%
    filter(n_retained_waves >= 2) %>%
    pull(SubNum)

  longitudinal_participants <- crosssec_data() %>%
    filter(SubNum %in% longitudinal_ids) %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_participants, longitudinal_participants) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group) %>%
    summarise(
      n = n(),
      Age_M = mean(Age_DuringParticipation_, na.rm = TRUE),
      Age_SD = sd(Age_DuringParticipation_, na.rm = TRUE),
      Education_M = mean(Education, na.rm = TRUE),
      Education_SD = sd(Education, na.rm = TRUE),
      Male_n = sum(Sex_M1 == 1, na.rm = TRUE),
      Female_n = sum(Sex_M1 == 2, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      `Age, M (SD)` = sprintf("%.2f (%.2f)", Age_M, Age_SD),
      `Education, M (SD)` = sprintf(
        "%.2f (%.2f)",
        Education_M,
        Education_SD
      ),
      `Male, n (%)` = sprintf("%d (%.1f%%)", Male_n, 100 * Male_n / n),
      `Female, n (%)` = sprintf(
        "%d (%.1f%%)",
        Female_n,
        100 * Female_n / n
      )
    ) %>%
    select(
      Sample,
      `Age group` = Age_Group,
      N = n,
      `Age, M (SD)`,
      `Education, M (SD)`,
      `Male, n (%)`,
      `Female, n (%)`
    )
})

descriptives_summary <- reactive({
  crosssec_descriptives <- crosssec_data() %>%
    group_by(SubNum, Age_Group, Modality) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(Sample = "Cross-sectional")

  baseline_age_groups <- crosssec_data() %>%
    distinct(SubNum, Age_Group)

  longitudinal_descriptives <- analysis_data() %>%
    group_by(SubNum, Modality) %>%
    filter(n_distinct(Wave_Num) >= 2) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    inner_join(baseline_age_groups, by = "SubNum") %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_descriptives, longitudinal_descriptives) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group, Modality) %>%
    summarise(
      Reliability_M = mean(Reliability, na.rm = TRUE),
      Reliability_SD = sd(Reliability, na.rm = TRUE),
      Confusability_M = mean(Confusability, na.rm = TRUE),
      Confusability_SD = sd(Confusability, na.rm = TRUE),
      Distinctiveness_M = mean(Distinctiveness, na.rm = TRUE),
      Distinctiveness_SD = sd(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      Modality = recode(
        as.character(Modality),
        Aud = "Auditory",
        Mot = "Motor",
        Vis = "Visual"
      )
    ) %>%
    arrange(
      Sample,
      Age_Group,
      factor(Modality, levels = c("Auditory", "Motor", "Visual"))
    )
})

style_apa_table <- function(ft) {
  ft %>%
    theme_booktabs() %>%
    font(fontname = "Times New Roman", part = "all") %>%
    fontsize(size = 11, part = "all") %>%
    bold(part = "header") %>%
    align(align = "center", part = "all") %>%
    align(j = 1, align = "left", part = "all") %>%
    padding(padding = 4, part = "all") %>%
    autofit()
}

demographics_flextable <- reactive({
  flextable(demographics_summary()) %>%
    merge_v(j = "Sample") %>%
    valign(j = "Sample", valign = "top") %>%
    style_apa_table()
})

descriptives_flextable <- reactive({
  flextable(descriptives_summary()) %>%
    colformat_double(
      j = c(
        "Reliability_M", "Reliability_SD",
        "Confusability_M", "Confusability_SD",
        "Distinctiveness_M", "Distinctiveness_SD"
      ),
      digits = 3,
      na_str = "—"
    ) %>%
    set_header_labels(
      Sample = "Sample",
      Age_Group = "Age group",
      Modality = "Modality",
      Reliability_M = "M",
      Reliability_SD = "SD",
      Confusability_M = "M",
      Confusability_SD = "SD",
      Distinctiveness_M = "M",
      Distinctiveness_SD = "SD"
    ) %>%
    add_header_row(
      values = c(
        "Sample", "Age group", "Modality", "Reliability",
        "Confusability", "Distinctiveness"
      ),
      colwidths = c(1, 1, 1, 2, 2, 2)
    ) %>%
    merge_v(j = "Sample") %>%
    merge_v(
      j = c("Sample", "Age_Group"),
      target = "Age_Group",
      combine = TRUE
    ) %>%
    valign(
      j = c("Sample", "Age_Group"),
      valign = "top"
    ) %>%
    style_apa_table()
})

apa_table_html <- function(number, title, note, ft) {
  htmltools::tagList(
    htmltools::tags$p(
      htmltools::tags$strong(number),
      style = "font-family: 'Times New Roman'; margin-bottom: 0;"
    ),
    htmltools::tags$p(
      htmltools::tags$em(title),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 1.05em; margin-top: 0;"
      )
    ),
    flextable::htmltools_value(ft),
    htmltools::tags$p(
      htmltools::tags$em("Note. "),
      paste0(note, " Selected rule: ", rule_label(), "."),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 0.95em; margin-top: 8px;"
      )
    )
  )
}
```

## Table 1

```{r demographics-table}
renderUI({
  apa_table_html(
    "Table 1",
    "Cross-Sectional and Longitudinal Analysis Demographics Summary",
    paste(
      "Values are presented as M (SD) unless otherwise indicated.",
      "The cross-sectional sample includes retained Wave 1 participants.",
      "The longitudinal sample includes participants with retained",
      "observations at Waves 1 and 2; their Wave 1 demographics are shown.",
      "Participants may therefore appear in both samples."
    ),
    demographics_flextable()
  )
})

downloadButton("download_demographics", "Download Table 1 (.docx)")

output$download_demographics <- downloadHandler(
  filename = function() "table1_demographics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 1. Cross-Sectional and Longitudinal Demographics" =
        demographics_flextable(),
      path = file
    )
  }
)
```

## Table 2

```{r descriptives-table}
renderUI({
  apa_table_html(
    "Table 2",
    "Cross-Sectional and Longitudinal Analysis Descriptive Statistics",
    paste(
      "Values are means and standard deviations for reliability",
      "(within-category similarity), confusability (between-category",
      "similarity), and distinctiveness by sample, age group, and modality.",
      "Longitudinal values are participant-level averages across two",
      "retained waves within each modality."
    ),
    descriptives_flextable()
  )
})

downloadButton("download_descriptives", "Download Table 2 (.docx)")

output$download_descriptives <- downloadHandler(
  filename = function() "table2_descriptive_statistics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 2. Cross-Sectional and Longitudinal Descriptives" =
        descriptives_flextable(),
      path = file
    )
  }
)
```

# Mixed-effects models

The fixed-effects specification is unchanged from the original document:
within-person age, between-person age, sex, education, and the two
effect-coded modality contrasts, with a participant random intercept.

```{r model-choice}
selectInput(
  "outcome",
  "Outcome shown below",
  choices = c("Reliability", "Confusability", "Distinctiveness"),
  selected = "Reliability"
)
```

## Model summary

```{r model-summary}
renderPrint({
  cat(rule_label(), "\n\n")
  print(summary(models()[[input$outcome]]))
})
```

## Type III tests

```{r model-anova}
renderPrint({
  cat(rule_label(), "\n\n")
  print(car::Anova(models()[[input$outcome]], type = 3))
})
```

# Outlier sensitivity (MAD x 3)

Outliers are flagged with a median absolute deviation criterion applied to each
outcome **separately within each modality**, because the auditory, motor, and
visual distributions differ enough that pooling them inflates the MAD and
flags almost nothing. An observation is flagged when
`abs(x - median(x)) > 3 * mad(x)`; R's `mad()` already applies the 1.4826
consistency constant.

Two exclusion variants are then compared with the full sample. A row is treated
as flagged if it is an outlier on *any* of the three outcomes, so all three
models are refit on the same subsample and the estimates stay comparable
across outcomes.

- **Drop flagged observations** removes only the flagged session x modality rows.
- **Drop flagged participants** removes every row belonging to a participant
  with at least one flagged observation.

The model specification is unchanged: `outcome ~ within_cAge + between_cAge +
Sex_M1 + Education + aud_mot + aud_vis + (1 | SubNum)`, `REML = FALSE`, with
Satterthwaite *p* values from `lmerTest`.

```{r outlier-sensitivity-helpers, include=FALSE}
add_mad_outlier_flags <- function(data) {
  data %>%
    group_by(Modality) %>%
    mutate(
      across(
        all_of(OUTCOMES),
        ~ !is.na(.x) &
          abs(.x - median(.x, na.rm = TRUE)) >
            MAD_THRESHOLD * mad(.x, na.rm = TRUE),
        .names = "outlier_{.col}"
      )
    ) %>%
    ungroup() %>%
    mutate(
      outlier_any = if_any(starts_with("outlier_"))
    )
}

mad_outlier_counts <- function(data) {
  flagged <- add_mad_outlier_flags(data)

  by_outcome <- map_dfr(OUTCOMES, function(outcome) {
    indicator <- flagged[[paste0("outlier_", outcome)]]
    tibble(
      Outcome = outcome,
      `Observations flagged` = sum(indicator),
      `Observations flagged (%)` = 100 * mean(indicator),
      `Participants affected` = n_distinct(flagged$SubNum[indicator])
    )
  })

  bind_rows(
    by_outcome,
    tibble(
      Outcome = "Any outcome",
      `Observations flagged` = sum(flagged$outlier_any),
      `Observations flagged (%)` = 100 * mean(flagged$outlier_any),
      `Participants affected` =
        n_distinct(flagged$SubNum[flagged$outlier_any])
    )
  )
}

build_outlier_variants <- function(data) {
  flagged <- add_mad_outlier_flags(data)
  affected_participants <- flagged %>%
    filter(outlier_any) %>%
    distinct(SubNum) %>%
    pull(SubNum)

  list(
    `Full sample` = flagged,
    `Drop flagged observations` = flagged %>% filter(!outlier_any),
    `Drop flagged participants` = flagged %>%
      filter(!SubNum %in% affected_participants)
  )
}

outlier_sensitivity_table <- function(data, rule_label_text = NA_character_) {
  variants <- build_outlier_variants(data)

  map_dfr(OUTCOMES, function(outcome) {
    map_dfr(names(variants), function(variant) {
      model <- fit_model(variants[[variant]], outcome)
      model_frame <- model@frame
      coefficients <- summary(model)$coefficients

      tibble(
        `Motion rule` = rule_label_text,
        Outcome = outcome,
        `Exclusion variant` = variant,
        `N observations` = nrow(model_frame),
        `N participants` = n_distinct(model_frame$SubNum),
        age_between_b = coefficients["between_cAge", "Estimate"],
        age_between_SE = coefficients["between_cAge", "Std. Error"],
        age_between_df = coefficients["between_cAge", "df"],
        age_between_p = coefficients["between_cAge", "Pr(>|t|)"],
        age_within_b = coefficients["within_cAge", "Estimate"],
        age_within_SE = coefficients["within_cAge", "Std. Error"],
        age_within_df = coefficients["within_cAge", "df"],
        age_within_p = coefficients["within_cAge", "Pr(>|t|)"]
      )
    })
  })
}

format_outlier_sensitivity <- function(table) {
  table %>%
    mutate(
      across(
        c(age_between_b, age_between_SE, age_within_b, age_within_SE),
        ~ sprintf("%.5f", .x)
      ),
      across(c(age_between_df, age_within_df), ~ sprintf("%.1f", .x)),
      across(c(age_between_p, age_within_p), format_p)
    ) %>%
    select(-any_of("Motion rule")) %>%
    rename(
      `age_between b` = age_between_b,
      `age_between SE` = age_between_SE,
      `age_between df` = age_between_df,
      `age_between p` = age_between_p,
      `age_within b` = age_within_b,
      `age_within SE` = age_within_SE,
      `age_within df` = age_within_df,
      `age_within p` = age_within_p
    )
}

outlier_sensitivity_by_rule <- lapply(MOTION_RULES, function(rule) {
  outlier_sensitivity_table(
    build_analysis_data(rule$fd_cutoff, rule$pct_cutoff),
    rule$label
  )
})
names(outlier_sensitivity_by_rule) <- vapply(
  MOTION_RULES, `[[`, character(1), "key"
)
```

## Flagged observations under the selected rule

```{r outlier-counts}
renderPrint({
  data <- analysis_data()
  counts <- mad_outlier_counts(data)
  cat(rule_label(), "\n")
  cat("Analytic observations:", nrow(data), "\n")
  cat("Participants:", n_distinct(data$SubNum), "\n\n")
  for (i in seq_len(nrow(counts))) {
    cat(sprintf(
      "%-16s %3d observations flagged (%.1f%%) across %d participants\n",
      counts$Outcome[[i]],
      counts$`Observations flagged`[[i]],
      counts$`Observations flagged (%)`[[i]],
      counts$`Participants affected`[[i]]
    ))
  }
})
```

## Comparison table under the selected rule

```{r outlier-sensitivity-table}
renderTable({
  format_outlier_sensitivity(outlier_sensitivity_table(analysis_data()))
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Comparison tables for both manuscript rules

These two tables do not depend on the controls above; they are the numbers
written to `outlier_sensitivity.csv`.

```{r outlier-sensitivity-both-rules}
renderUI({
  htmltools::tagList(lapply(MOTION_RULES, function(rule) {
    table <- outlier_sensitivity_by_rule[[rule$key]]
    htmltools::tagList(
      htmltools::tags$h4(rule$label),
      htmltools::HTML(
        knitr::kable(
          format_outlier_sensitivity(table),
          format = "html",
          row.names = FALSE,
          table.attr = "class='table table-striped table-bordered'"
        )
      )
    )
  }))
})
```

# Exported numbers for the manuscript

```{r exports, include=FALSE}
sample_definitions <- function(data) {
  returner_ids <- data %>%
    distinct(SubNum, Wave_Num) %>%
    count(SubNum, name = "n_retained_waves") %>%
    filter(n_retained_waves >= 2) %>%
    pull(SubNum)

  young_wave1 <- data %>%
    filter(Wave_Num == 1, Age_DuringParticipation_ < OLDER_ADULT_AGE)
  older_wave1 <- data %>%
    filter(Wave_Num == 1, Age_DuringParticipation_ >= OLDER_ADULT_AGE)

  list(
    `Young adults, Wave 1` = young_wave1,
    `Older adults, Wave 1` = older_wave1,
    `Older adults, Wave 2 (returners)` = data %>%
      filter(
        Wave_Num == 2,
        Age_DuringParticipation_ >= OLDER_ADULT_AGE,
        SubNum %in% returner_ids
      ),
    `Older adults, Wave 1 (Wave 2 returners)` = older_wave1 %>%
      filter(SubNum %in% returner_ids)
  )
}

demographics_export <- function(data, rule_label_text) {
  samples <- sample_definitions(data)

  map_dfr(names(samples), function(sample_name) {
    participants <- samples[[sample_name]] %>%
      distinct(SubNum, .keep_all = TRUE)

    tibble(
      motion_rule = rule_label_text,
      sample = sample_name,
      n = nrow(participants),
      age_m = mean(participants$Age_DuringParticipation_, na.rm = TRUE),
      age_sd = sd(participants$Age_DuringParticipation_, na.rm = TRUE),
      age_min = min(participants$Age_DuringParticipation_, na.rm = TRUE),
      age_max = max(participants$Age_DuringParticipation_, na.rm = TRUE),
      education_m = mean(participants$Education, na.rm = TRUE),
      education_sd = sd(participants$Education, na.rm = TRUE),
      n_men = sum(participants$Sex_M1 == 1, na.rm = TRUE),
      n_women = sum(participants$Sex_M1 == 2, na.rm = TRUE)
    )
  })
}

descriptives_export <- function(data, rule_label_text) {
  samples <- sample_definitions(data)

  map_dfr(names(samples), function(sample_name) {
    samples[[sample_name]] %>%
      mutate(
        modality = recode(
          as.character(Modality),
          Aud = "Auditory",
          Mot = "Motor",
          Vis = "Visual"
        )
      ) %>%
      group_by(modality) %>%
      summarise(
        n = sum(!is.na(Reliability)),
        reliability_m = mean(Reliability, na.rm = TRUE),
        reliability_sd = sd(Reliability, na.rm = TRUE),
        confusability_m = mean(Confusability, na.rm = TRUE),
        confusability_sd = sd(Confusability, na.rm = TRUE),
        distinctiveness_m = mean(Distinctiveness, na.rm = TRUE),
        distinctiveness_sd = sd(Distinctiveness, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        motion_rule = rule_label_text,
        sample = sample_name,
        .before = 1
      )
  })
}

motion_accounting_export <- function() {
  map_dfr(MOTION_RULES, function(rule) {
    flags <- build_run_flags(rule$fd_cutoff, rule$pct_cutoff) %>%
      left_join(subject_age_groups, by = "Subject") %>%
      mutate(
        age_group = as.character(Age_Group),
        modality = recode(
          as.character(Modality),
          Aud = "Auditory",
          Mot = "Motor",
          Vis = "Visual"
        )
      )

    summarise_flags <- function(grouped) {
      grouped %>%
        summarise(
          runs_assessed = n(),
          runs_excluded = sum(exclude, na.rm = TRUE),
          pct_runs_excluded = 100 * mean(exclude, na.rm = TRUE),
          mean_pct_volumes_scrubbed = mean(pct_excluded, na.rm = TRUE),
          .groups = "drop"
        )
    }

    bind_rows(
      summarise_flags(flags %>% group_by(age_group, modality)),
      summarise_flags(flags %>% group_by(age_group)) %>%
        mutate(modality = "All modalities"),
      summarise_flags(flags %>% group_by(modality)) %>%
        mutate(age_group = "All participants"),
      summarise_flags(flags %>% group_by()) %>%
        mutate(age_group = "All participants", modality = "All modalities")
    ) %>%
      mutate(motion_rule = rule$label, .before = 1) %>%
      select(
        motion_rule, age_group, modality, runs_assessed, runs_excluded,
        pct_runs_excluded, mean_pct_volumes_scrubbed
      )
  })
}

model_estimates_export <- function() {
  map_dfr(MOTION_RULES, function(rule) {
    data <- build_analysis_data(rule$fd_cutoff, rule$pct_cutoff)
    outcome_models <- fit_outcome_models(data)

    map_dfr(OUTCOMES, function(outcome) {
      model <- outcome_models[[outcome]]
      summary(model)$coefficients %>%
        as.data.frame() %>%
        rownames_to_column("term") %>%
        as_tibble() %>%
        transmute(
          motion_rule = rule$label,
          outcome = outcome,
          term,
          estimate = Estimate,
          se = `Std. Error`,
          df,
          t = `t value`,
          p = `Pr(>|t|)`,
          n_observations = nrow(model@frame),
          n_participants = n_distinct(model@frame$SubNum)
        )
    })
  })
}

export_tables <- list()
for (rule in MOTION_RULES) {
  rule_data <- build_analysis_data(rule$fd_cutoff, rule$pct_cutoff)
  export_tables[[paste0(rule$prefix, "_demographics.csv")]] <-
    demographics_export(rule_data, rule$label)
  export_tables[[paste0(rule$prefix, "_descriptives.csv")]] <-
    descriptives_export(rule_data, rule$label)
}
export_tables[["motion_accounting.csv"]] <- motion_accounting_export()
export_tables[["model_estimates.csv"]] <- model_estimates_export()
export_tables[["outlier_sensitivity.csv"]] <-
  bind_rows(outlier_sensitivity_by_rule)

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
for (filename in names(export_tables)) {
  readr::write_csv(
    export_tables[[filename]],
    file.path(output_dir, filename)
  )
}
```

The tables the manuscript needs are written to `output/` when this document
starts. `demographics.csv` and `descriptives.csv` are rule-specific and are
therefore prefixed with the motion rule; the remaining files carry a
`motion_rule` column and cover both rules in one file.

```{r export-summary}
renderTable({
  tibble(
    File = file.path("output", names(export_tables)),
    Rows = vapply(export_tables, nrow, integer(1)),
    Columns = vapply(export_tables, ncol, integer(1))
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# Interactive age-effect figure

The combined figure contains auditory, motor, and visual ROI panels. The blue
and red lines are the model-implied cross-sectional and longitudinal age
effects from ROI-specific mixed-effects models adjusted for sex and education.
Each panel reports the corresponding unstandardized coefficient (*B*),
standardized coefficient (β), and *p* value.

```{r plot-helper, include=FALSE}
format_age_p <- function(p) {
  if (is.na(p)) "NA" else if (p < 0.001) "< .001" else sprintf("%.3f", p)
}

get_roi_age_plot_info <- function(data, modality, outcome) {
  plot_data <- data %>%
    filter(Modality == modality) %>%
    select(
      Subject, SubNum, Wave_Num, Age_DuringParticipation_, n_waves,
      mean_age, within_cAge, between_cAge, Sex_M1, Education,
      Modality, all_of(outcome)
    ) %>%
    rename(value = all_of(outcome)) %>%
    drop_na(
      value, Age_DuringParticipation_, within_cAge, between_cAge,
      Sex_M1, Education, SubNum
    )

  roi_model <- lmer(
    reformulate(
      c(
        "within_cAge", "between_cAge", "Sex_M1", "Education",
        "(1 | SubNum)"
      ),
      response = "value"
    ),
    data = plot_data,
    REML = FALSE
  )

  coefficients <- summary(roi_model)$coefficients
  outcome_sd <- sd(plot_data$value, na.rm = TRUE)
  within_b <- coefficients["within_cAge", "Estimate"]
  between_b <- coefficients["between_cAge", "Estimate"]
  within_beta <- within_b * sd(plot_data$within_cAge, na.rm = TRUE) / outcome_sd
  between_beta <- between_b * sd(plot_data$between_cAge, na.rm = TRUE) / outcome_sd
  stats_label <- sprintf(
    paste0(
      "Long.: B=%.3f, β=%.3f, p=%s\n",
      "Cross-sec.: B=%.3f, β=%.3f, p=%s"
    ),
    within_b,
    within_beta,
    format_age_p(coefficients["within_cAge", "Pr(>|t|)"]),
    between_b,
    between_beta,
    format_age_p(coefficients["between_cAge", "Pr(>|t|)"])
  )

  longitudinal_data <- plot_data %>% filter(n_waves > 1)
  typical_sex <- median(plot_data$Sex_M1, na.rm = TRUE)
  typical_education <- median(plot_data$Education, na.rm = TRUE)

  between_age <- seq(
    min(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    max(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    length.out = 100
  )
  between_line <- tibble(
    age = between_age,
    within_cAge = 0,
    between_cAge = between_age - OLDER_ADULT_AGE,
    Sex_M1 = typical_sex,
    Education = typical_education,
    SubNum = plot_data$SubNum[1]
  ) %>%
    mutate(predicted = predict(
      roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
    ))

  if (nrow(longitudinal_data) > 0 &&
      diff(range(longitudinal_data$within_cAge, na.rm = TRUE)) > 0) {
    within_values <- seq(
      min(longitudinal_data$within_cAge, na.rm = TRUE),
      max(longitudinal_data$within_cAge, na.rm = TRUE),
      length.out = 100
    )
    within_age <- mean(longitudinal_data$mean_age, na.rm = TRUE) +
      within_values
    within_line <- tibble(
      age = within_age,
      within_cAge = within_values,
      between_cAge = 0,
      Sex_M1 = typical_sex,
      Education = typical_education,
      SubNum = plot_data$SubNum[1]
    ) %>%
      mutate(predicted = predict(
        roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
      ))
  } else {
    within_line <- tibble(age = numeric(), predicted = numeric())
  }

  list(
    data = plot_data,
    longitudinal_data = longitudinal_data,
    between_line = between_line,
    within_line = within_line,
    stats_label = stats_label
  )
}

make_roi_age_panel <- function(data, modality, outcome, label) {
  plot_info <- get_roi_age_plot_info(data, modality, outcome)
  roi_labels <- c(Aud = "Auditory", Mot = "Motor", Vis = "Visual")

  point_data <- plot_info$data

  figure <- ggplot(point_data, aes(Age_DuringParticipation_, value)) +
    geom_line(
      data = plot_info$longitudinal_data %>% arrange(SubNum, Wave_Num),
      aes(group = SubNum),
      colour = "#CC0000", linewidth = 0.35, alpha = 0.45
    ) +
    geom_point(size = 1.25, alpha = 0.5) +
    geom_line(
      data = plot_info$between_line,
      aes(age, predicted, colour = "Cross-sectional"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = plot_info$within_line,
      aes(age, predicted, colour = "Longitudinal"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    scale_colour_manual(
      values = c("Cross-sectional" = "#3366CC", "Longitudinal" = "#CC0000"),
      name = NULL
    ) +
    labs(
      title = unname(roi_labels[[modality]]),
      subtitle = plot_info$stats_label,
      x = "Age",
      y = paste("Neural", tolower(outcome)),
      caption = label
    ) +
    theme_classic(base_size = 10) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 6.8, hjust = 0.5, lineheight = 1.1),
      plot.caption = element_text(size = 8.5, hjust = 0.5),
      axis.title = element_text(size = 9.5),
      axis.text = element_text(size = 8.5),
      legend.text = element_text(size = 8.5)
    )

  list(
    figure = figure,
    roi_label = unname(roi_labels[[modality]]),
    stats_label = plot_info$stats_label
  )
}
```

```{r age-figure, fig.width=15, fig.height=5.2, out.width='100%'}
renderPlot({
  req(input$outcome)

  modalities <- c("Aud", "Mot", "Vis")
  panels <- lapply(
    modalities,
    function(modality) {
      make_roi_age_panel(
        analysis_data(),
        modality,
        input$outcome,
        ""
      )
    }
  )

  wrap_plots(lapply(panels, `[[`, "figure"), nrow = 1) +
    plot_layout(guides = "collect") +
    plot_annotation(
      title = paste(input$outcome, "age effects"),
      subtitle = rule_label(),
      theme = theme(
        plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 10)
      )
    ) &
    theme(legend.position = "bottom")
}, res = 110)
```

# Excluded runs

```{r excluded-runs}
renderTable({
  selected_flags() %>%
    filter(exclude) %>%
    arrange(desc(pct_excluded), Subject, Modality) %>%
    select(Subject, Modality, pct_excluded) %>%
    rename(`Percent excluded TRs` = pct_excluded)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```
