---
title: "Motion-Corrected Analysis (Interactive)"
author: "Tien"
date: "2026-07-26"
output:
  html_document:
    toc: true
    toc_float: true
runtime: shiny
---

```{r setup, include=FALSE}
required_packages <- c(
  "shiny", "tidyverse", "lme4", "lmerTest", "car", "patchwork",
  "flextable"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install these packages before running the document: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(shiny)
library(tidyverse)
library(lme4)
library(lmerTest)
library(patchwork)
library(flextable)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

<style>
.main-container {
  max-width: 1500px !important;
}
.shiny-plot-output {
  width: 100% !important;
}
</style>

# Choose the motion-exclusion rule

Choose the framewise-displacement (FD) cutoff and the percentage of excluded
TRs that causes a run to be removed. The models, sample summary, and plots
below update automatically.

```{r controls}
fluidRow(
  column(
    width = 5,
    selectInput(
      "fd_cutoff",
      "FD cutoff",
      choices = c("0.5 mm" = "0.5", "0.4 mm" = "0.4"),
      selected = "0.5"
    )
  ),
  column(
    width = 7,
    sliderInput(
      "pct_cutoff",
      "Exclude a run when percentage of excluded TRs is at least:",
      min = 0,
      max = 100,
      value = 20,
      step = 1,
      post = "%"
    )
  )
)

helpText(
  "The previous liberal rule was 0.5 mm / 20%; ",
  "the previous conservative rule was 0.4 mm / 10%."
)

observeEvent(input$fd_cutoff, {
  updateSliderInput(
    session,
    "pct_cutoff",
    value = if (identical(input$fd_cutoff, "0.4")) 10 else 20
  )
}, ignoreInit = TRUE)
```

# Data preparation

```{r load-and-prepare, include=FALSE}
project_dir <- Sys.getenv("MIND_PROJECT_DIR", unset = NA_character_)
if (is.na(project_dir) || !dir.exists(project_dir)) {
  stop(
    "The MiND project directory is not set. Launch this document using ",
    "'Run Motion Corrected Analysis Interactive.R'."
  )
}
project_dir <- normalizePath(project_dir, mustWork = TRUE)
motion_dir <- file.path(project_dir, "Data", "Motion")

master_data_part1 <- readr::read_csv(
  file.path(motion_dir, "CombinedMasterGLX_new_part1.csv"),
  num_threads = 1,
  show_col_types = FALSE
)
master_data_part2 <- readr::read_csv(
  file.path(motion_dir, "CombinedMasterGLX_new_part2.csv"),
  num_threads = 1,
  show_col_types = FALSE
)

motion_files <- c(
  "0.5" = "motion_exclusion_by_run_5mm.csv",
  "0.4" = "motion_exclusion_by_run_4mm.csv"
)
motion_exclusions <- lapply(motion_files, function(filename) {
  readr::read_csv(
    file.path(motion_dir, filename),
    num_threads = 1,
    show_col_types = FALSE
  )
})

df_mind <- master_data_part2 %>%
  select(
    Subject,
    D_aud_sm_sm2000_MC_P, D_vis_sm_sm2000_MC_P, D_mot_sm_sm2000_MC_P,
    W_aud_sm_sm2000_MC_P, W_vis_sm_sm2000_MC_P, W_mot_sm_sm2000_MC_P,
    B_aud_sm_sm2000_MC_P, B_vis_sm_sm2000_MC_P, B_mot_sm_sm2000_MC_P
  ) %>%
  rename(
    AudDistinctiveness = D_aud_sm_sm2000_MC_P,
    MotDistinctiveness = D_mot_sm_sm2000_MC_P,
    VisDistinctiveness = D_vis_sm_sm2000_MC_P,
    AudReliability = W_aud_sm_sm2000_MC_P,
    MotReliability = W_mot_sm_sm2000_MC_P,
    VisReliability = W_vis_sm_sm2000_MC_P,
    AudConfusability = B_aud_sm_sm2000_MC_P,
    MotConfusability = B_mot_sm_sm2000_MC_P,
    VisConfusability = B_vis_sm_sm2000_MC_P
  ) %>%
  filter(!is.na(Subject))

df_mind_mc04 <- master_data_part2 %>%
  select(
    Subject,
    D_aud_sm_sm2000_MC04_P, D_vis_sm_sm2000_MC04_P, D_mot_sm_sm2000_MC04_P,
    W_aud_sm_sm2000_MC04_P, W_vis_sm_sm2000_MC04_P, W_mot_sm_sm2000_MC04_P,
    B_aud_sm_sm2000_MC04_P, B_vis_sm_sm2000_MC04_P, B_mot_sm_sm2000_MC04_P
  ) %>%
  rename(
    AudDistinctiveness = D_aud_sm_sm2000_MC04_P,
    MotDistinctiveness = D_mot_sm_sm2000_MC04_P,
    VisDistinctiveness = D_vis_sm_sm2000_MC04_P,
    AudReliability = W_aud_sm_sm2000_MC04_P,
    MotReliability = W_mot_sm_sm2000_MC04_P,
    VisReliability = W_vis_sm_sm2000_MC04_P,
    AudConfusability = B_aud_sm_sm2000_MC04_P,
    MotConfusability = B_mot_sm_sm2000_MC04_P,
    VisConfusability = B_vis_sm_sm2000_MC04_P
  ) %>%
  filter(!is.na(Subject))

df_mind_demo <- master_data_part1 %>%
  select(
    Subject, SubNum, Age_DuringParticipation_, Subgroup, Wave_Num,
    Sex_M1, Education
  )

prepare_outcome_data <- function(data) {
  data %>%
    left_join(df_mind_demo, by = "Subject") %>%
    filter(!is.na(SubNum)) %>%
    pivot_longer(
      cols = matches("^(Aud|Mot|Vis)(Distinctiveness|Reliability|Confusability)$"),
      names_to = c("Modality", ".value"),
      names_pattern = "(Aud|Mot|Vis)(.*)"
    ) %>%
    mutate(Modality = factor(Modality, levels = c("Aud", "Mot", "Vis")))
}

df_mind_long <- prepare_outcome_data(df_mind)
df_mind_long_mc04 <- prepare_outcome_data(df_mind_mc04)

outcome_data <- list(
  "0.5" = df_mind_long,
  "0.4" = df_mind_long_mc04
)

healthy_subject_ids <- bind_rows(df_mind_long, df_mind_long_mc04) %>%
  filter(Subgroup == 1) %>%
  distinct(Subject) %>%
  pull(Subject)
```

```{r reactive-analysis, include=FALSE}
selected_flags <- reactive({
  req(input$fd_cutoff, input$pct_cutoff)

  motion_exclusions[[input$fd_cutoff]] %>%
    filter(
      subject %in% healthy_subject_ids,
      session == "placebo",
      run == "001",
      experiment %in% c("auditory", "visual", "motor")
    ) %>%
    transmute(
      Subject = subject,
      Modality = recode(
        experiment,
        auditory = "Aud",
        motor = "Mot",
        visual = "Vis"
      ),
      pct_excluded,
      exclude = pct_excluded >= input$pct_cutoff
    )
})

motion_filtered_data <- reactive({
  req(input$fd_cutoff)

  outcome_data[[input$fd_cutoff]] %>%
    left_join(selected_flags(), by = c("Subject", "Modality")) %>%
    # Match the non-interactive report: rows without a motion record are dropped.
    filter(!exclude) %>%
    filter(Wave_Num %in% c(1, 2, 3))
})

add_age_decomposition <- function(data) {
  data %>%
    group_by(SubNum) %>%
    mutate(
      n_waves = n_distinct(Wave_Num),
      mean_age = if_else(
        n_waves > 1,
        mean(Age_DuringParticipation_, na.rm = TRUE),
        Age_DuringParticipation_
      ),
      within_cAge = if_else(
        n_waves > 1,
        Age_DuringParticipation_ - mean_age,
        0
      ),
      between_cAge = mean_age - 65,
      aud_mot = case_when(
        Modality == "Aud" ~ -1,
        Modality == "Mot" ~ 1,
        TRUE ~ 0
      ),
      aud_vis = case_when(
        Modality == "Aud" ~ -1,
        Modality == "Vis" ~ 1,
        TRUE ~ 0
      )
    ) %>%
    ungroup()
}

analysis_data <- reactive({
  motion_filtered_data() %>%
    filter(Subgroup == 1, Wave_Num %in% c(1, 2)) %>%
    add_age_decomposition()
})

fit_model <- function(data, outcome) {
  model_data <- data %>%
    drop_na(
      all_of(c(
        outcome, "within_cAge", "between_cAge", "Sex_M1", "Education",
        "aud_mot", "aud_vis", "SubNum"
      ))
    )

  validate(need(
    nrow(model_data) > 0,
    paste("No complete observations are available for", outcome)
  ))

  lmer(
    reformulate(
      c(
        "within_cAge", "between_cAge", "Sex_M1", "Education",
        "aud_mot", "aud_vis", "(1 | SubNum)"
      ),
      response = outcome
    ),
    data = model_data,
    REML = FALSE
  )
}

models <- reactive({
  data <- analysis_data()
  setNames(
    lapply(
      c("Reliability", "Confusability", "Distinctiveness"),
      function(outcome) fit_model(data, outcome)
    ),
    c("Reliability", "Confusability", "Distinctiveness")
  )
})

rule_label <- reactive({
  paste0(
    "FD ≥ ", input$fd_cutoff, " mm; run excluded at ≥ ",
    input$pct_cutoff, "% excluded TRs"
  )
})
```

# Sample after exclusion

```{r sample-summary}
renderTable({
  flags <- selected_flags()
  kept <- analysis_data()

  tibble(
    `Selected rule` = rule_label(),
    `Runs flagged` = sum(flags$exclude, na.rm = TRUE),
    `Runs assessed` = nrow(flags),
    `Unique scans retained` = n_distinct(kept$Subject),
    `Unique participants retained` = n_distinct(kept$SubNum),
    `Modality-level observations retained` = nrow(kept)
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

# APA-style demographics and descriptive statistics

These tables use the currently selected motion-exclusion rule. The
cross-sectional sample is defined as Wave 1. The longitudinal sample requires
retained observations at both Waves 1 and 2; Wave 1 demographics are reported
so each longitudinal participant is counted once. Young adults are below age
65 and older adults are age 65 or above. In Table 2, longitudinal descriptive
statistics average each participant's two retained waves within modality
before calculating the group mean and standard deviation.

```{r apa-table-data, include=FALSE}
crosssec_data <- reactive({
  analysis_data() %>%
    filter(Wave_Num == 1) %>%
    mutate(
      Age_Group = factor(
        if_else(
          Age_DuringParticipation_ < 65,
          "Young adults",
          "Older adults"
        ),
        levels = c("Young adults", "Older adults")
      )
    )
})

demographics_summary <- reactive({
  crosssec_participants <- crosssec_data() %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Cross-sectional")

  longitudinal_ids <- analysis_data() %>%
    distinct(SubNum, Wave_Num) %>%
    count(SubNum, name = "n_retained_waves") %>%
    filter(n_retained_waves >= 2) %>%
    pull(SubNum)

  longitudinal_participants <- crosssec_data() %>%
    filter(SubNum %in% longitudinal_ids) %>%
    distinct(SubNum, .keep_all = TRUE) %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_participants, longitudinal_participants) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group) %>%
    summarise(
      n = n(),
      Age_M = mean(Age_DuringParticipation_, na.rm = TRUE),
      Age_SD = sd(Age_DuringParticipation_, na.rm = TRUE),
      Education_M = mean(Education, na.rm = TRUE),
      Education_SD = sd(Education, na.rm = TRUE),
      Male_n = sum(Sex_M1 == 1, na.rm = TRUE),
      Female_n = sum(Sex_M1 == 2, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      `Age, M (SD)` = sprintf("%.2f (%.2f)", Age_M, Age_SD),
      `Education, M (SD)` = sprintf(
        "%.2f (%.2f)",
        Education_M,
        Education_SD
      ),
      `Male, n (%)` = sprintf("%d (%.1f%%)", Male_n, 100 * Male_n / n),
      `Female, n (%)` = sprintf(
        "%d (%.1f%%)",
        Female_n,
        100 * Female_n / n
      )
    ) %>%
    select(
      Sample,
      `Age group` = Age_Group,
      N = n,
      `Age, M (SD)`,
      `Education, M (SD)`,
      `Male, n (%)`,
      `Female, n (%)`
    )
})

descriptives_summary <- reactive({
  crosssec_descriptives <- crosssec_data() %>%
    group_by(SubNum, Age_Group, Modality) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(Sample = "Cross-sectional")

  baseline_age_groups <- crosssec_data() %>%
    distinct(SubNum, Age_Group)

  longitudinal_descriptives <- analysis_data() %>%
    group_by(SubNum, Modality) %>%
    filter(n_distinct(Wave_Num) >= 2) %>%
    summarise(
      Reliability = mean(Reliability, na.rm = TRUE),
      Confusability = mean(Confusability, na.rm = TRUE),
      Distinctiveness = mean(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    inner_join(baseline_age_groups, by = "SubNum") %>%
    mutate(Sample = "Longitudinal")

  bind_rows(crosssec_descriptives, longitudinal_descriptives) %>%
    mutate(
      Sample = factor(
        Sample,
        levels = c("Cross-sectional", "Longitudinal")
      )
    ) %>%
    group_by(Sample, Age_Group, Modality) %>%
    summarise(
      Reliability_M = mean(Reliability, na.rm = TRUE),
      Reliability_SD = sd(Reliability, na.rm = TRUE),
      Confusability_M = mean(Confusability, na.rm = TRUE),
      Confusability_SD = sd(Confusability, na.rm = TRUE),
      Distinctiveness_M = mean(Distinctiveness, na.rm = TRUE),
      Distinctiveness_SD = sd(Distinctiveness, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      Modality = recode(
        as.character(Modality),
        Aud = "Auditory",
        Mot = "Motor",
        Vis = "Visual"
      )
    ) %>%
    arrange(
      Sample,
      Age_Group,
      factor(Modality, levels = c("Auditory", "Motor", "Visual"))
    )
})

style_apa_table <- function(ft) {
  ft %>%
    theme_booktabs() %>%
    font(fontname = "Times New Roman", part = "all") %>%
    fontsize(size = 11, part = "all") %>%
    bold(part = "header") %>%
    align(align = "center", part = "all") %>%
    align(j = 1, align = "left", part = "all") %>%
    padding(padding = 4, part = "all") %>%
    autofit()
}

demographics_flextable <- reactive({
  flextable(demographics_summary()) %>%
    merge_v(j = "Sample") %>%
    valign(j = "Sample", valign = "top") %>%
    style_apa_table()
})

descriptives_flextable <- reactive({
  flextable(descriptives_summary()) %>%
    colformat_double(
      j = c(
        "Reliability_M", "Reliability_SD",
        "Confusability_M", "Confusability_SD",
        "Distinctiveness_M", "Distinctiveness_SD"
      ),
      digits = 3,
      na_str = "\u2014"
    ) %>%
    set_header_labels(
      Sample = "Sample",
      Age_Group = "Age group",
      Modality = "Modality",
      Reliability_M = "M",
      Reliability_SD = "SD",
      Confusability_M = "M",
      Confusability_SD = "SD",
      Distinctiveness_M = "M",
      Distinctiveness_SD = "SD"
    ) %>%
    add_header_row(
      values = c(
        "Sample", "Age group", "Modality", "Reliability",
        "Confusability", "Distinctiveness"
      ),
      colwidths = c(1, 1, 1, 2, 2, 2)
    ) %>%
    merge_v(j = "Sample") %>%
    merge_v(
      j = c("Sample", "Age_Group"),
      target = "Age_Group",
      combine = TRUE
    ) %>%
    valign(
      j = c("Sample", "Age_Group"),
      valign = "top"
    ) %>%
    style_apa_table()
})

apa_table_html <- function(number, title, note, ft) {
  htmltools::tagList(
    htmltools::tags$p(
      htmltools::tags$strong(number),
      style = "font-family: 'Times New Roman'; margin-bottom: 0;"
    ),
    htmltools::tags$p(
      htmltools::tags$em(title),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 1.05em; margin-top: 0;"
      )
    ),
    flextable::htmltools_value(ft),
    htmltools::tags$p(
      htmltools::tags$em("Note. "),
      paste0(note, " Selected rule: ", rule_label(), "."),
      style = paste(
        "font-family: 'Times New Roman';",
        "font-size: 0.95em; margin-top: 8px;"
      )
    )
  )
}
```

## Table 1

```{r demographics-table}
renderUI({
  apa_table_html(
    "Table 1",
    "Cross-Sectional and Longitudinal Analysis Demographics Summary",
    paste(
      "Values are presented as M (SD) unless otherwise indicated.",
      "The cross-sectional sample includes retained Wave 1 participants.",
      "The longitudinal sample includes participants with retained",
      "observations at Waves 1 and 2; their Wave 1 demographics are shown.",
      "Participants may therefore appear in both samples."
    ),
    demographics_flextable()
  )
})

downloadButton("download_demographics", "Download Table 1 (.docx)")

output$download_demographics <- downloadHandler(
  filename = function() "table1_demographics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 1. Cross-Sectional and Longitudinal Demographics" =
        demographics_flextable(),
      path = file
    )
  }
)
```

## Table 2

```{r descriptives-table}
renderUI({
  apa_table_html(
    "Table 2",
    "Cross-Sectional and Longitudinal Analysis Descriptive Statistics",
    paste(
      "Values are means and standard deviations for reliability",
      "(within-category similarity), confusability (between-category",
      "similarity), and distinctiveness by sample, age group, and modality.",
      "Longitudinal values are participant-level averages across two",
      "retained waves within each modality."
    ),
    descriptives_flextable()
  )
})

downloadButton("download_descriptives", "Download Table 2 (.docx)")

output$download_descriptives <- downloadHandler(
  filename = function() "table2_descriptive_statistics.docx",
  content = function(file) {
    flextable::save_as_docx(
      "Table 2. Cross-Sectional and Longitudinal Descriptives" =
        descriptives_flextable(),
      path = file
    )
  }
)
```

# Mixed-effects models

The fixed-effects specification is unchanged from the original document:
within-person age, between-person age, sex, education, and the two
effect-coded modality contrasts, with a participant random intercept.

```{r model-choice}
selectInput(
  "outcome",
  "Outcome shown below",
  choices = c("Reliability", "Confusability", "Distinctiveness"),
  selected = "Reliability"
)
```

## Model summary

```{r model-summary}
renderPrint({
  cat(rule_label(), "\n\n")
  print(summary(models()[[input$outcome]]))
})
```

## Type III tests

```{r model-anova}
renderPrint({
  cat(rule_label(), "\n\n")
  print(car::Anova(models()[[input$outcome]], type = 3))
})
```

# Interactive age-effect figure

The combined figure contains auditory, motor, and visual ROI panels. The blue
and red lines are the model-implied cross-sectional and longitudinal age
effects from ROI-specific mixed-effects models adjusted for sex and education.
Each panel reports the corresponding unstandardized coefficient (*B*),
standardized coefficient (β), and *p* value.

```{r plot-helper, include=FALSE}
format_age_p <- function(p) {
  if (is.na(p)) "NA" else if (p < 0.001) "< .001" else sprintf("%.3f", p)
}

get_roi_age_plot_info <- function(data, modality, outcome) {
  plot_data <- data %>%
    filter(Modality == modality) %>%
    select(
      Subject, SubNum, Wave_Num, Age_DuringParticipation_, n_waves,
      mean_age, within_cAge, between_cAge, Sex_M1, Education,
      Modality, all_of(outcome)
    ) %>%
    rename(value = all_of(outcome)) %>%
    drop_na(
      value, Age_DuringParticipation_, within_cAge, between_cAge,
      Sex_M1, Education, SubNum
    )

  roi_model <- lmer(
    reformulate(
      c(
        "within_cAge", "between_cAge", "Sex_M1", "Education",
        "(1 | SubNum)"
      ),
      response = "value"
    ),
    data = plot_data,
    REML = FALSE
  )

  coefficients <- summary(roi_model)$coefficients
  outcome_sd <- sd(plot_data$value, na.rm = TRUE)
  within_b <- coefficients["within_cAge", "Estimate"]
  between_b <- coefficients["between_cAge", "Estimate"]
  within_beta <- within_b * sd(plot_data$within_cAge, na.rm = TRUE) / outcome_sd
  between_beta <- between_b * sd(plot_data$between_cAge, na.rm = TRUE) / outcome_sd
  stats_label <- sprintf(
    paste0(
      "Long.: B=%.3f, β=%.3f, p=%s\n",
      "Cross-sec.: B=%.3f, β=%.3f, p=%s"
    ),
    within_b,
    within_beta,
    format_age_p(coefficients["within_cAge", "Pr(>|t|)"]),
    between_b,
    between_beta,
    format_age_p(coefficients["between_cAge", "Pr(>|t|)"])
  )

  longitudinal_data <- plot_data %>% filter(n_waves > 1)
  typical_sex <- median(plot_data$Sex_M1, na.rm = TRUE)
  typical_education <- median(plot_data$Education, na.rm = TRUE)

  between_age <- seq(
    min(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    max(plot_data$Age_DuringParticipation_, na.rm = TRUE),
    length.out = 100
  )
  between_line <- tibble(
    age = between_age,
    within_cAge = 0,
    between_cAge = between_age - 65,
    Sex_M1 = typical_sex,
    Education = typical_education,
    SubNum = plot_data$SubNum[1]
  ) %>%
    mutate(predicted = predict(
      roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
    ))

  if (nrow(longitudinal_data) > 0 &&
      diff(range(longitudinal_data$within_cAge, na.rm = TRUE)) > 0) {
    within_values <- seq(
      min(longitudinal_data$within_cAge, na.rm = TRUE),
      max(longitudinal_data$within_cAge, na.rm = TRUE),
      length.out = 100
    )
    within_age <- mean(longitudinal_data$mean_age, na.rm = TRUE) +
      within_values
    within_line <- tibble(
      age = within_age,
      within_cAge = within_values,
      between_cAge = 0,
      Sex_M1 = typical_sex,
      Education = typical_education,
      SubNum = plot_data$SubNum[1]
    ) %>%
      mutate(predicted = predict(
        roi_model, newdata = ., re.form = NA, allow.new.levels = TRUE
      ))
  } else {
    within_line <- tibble(age = numeric(), predicted = numeric())
  }

  list(
    data = plot_data,
    longitudinal_data = longitudinal_data,
    between_line = between_line,
    within_line = within_line,
    stats_label = stats_label
  )
}

make_roi_age_panel <- function(data, modality, outcome, label) {
  plot_info <- get_roi_age_plot_info(data, modality, outcome)
  roi_labels <- c(Aud = "Auditory", Mot = "Motor", Vis = "Visual")

  point_data <- plot_info$data

  figure <- ggplot(point_data, aes(Age_DuringParticipation_, value)) +
    geom_line(
      data = plot_info$longitudinal_data %>% arrange(SubNum, Wave_Num),
      aes(group = SubNum),
      colour = "#CC0000", linewidth = 0.35, alpha = 0.45
    ) +
    geom_point(size = 1.25, alpha = 0.5) +
    geom_line(
      data = plot_info$between_line,
      aes(age, predicted, colour = "Cross-sectional"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = plot_info$within_line,
      aes(age, predicted, colour = "Longitudinal"),
      linewidth = 1.4,
      inherit.aes = FALSE
    ) +
    scale_colour_manual(
      values = c("Cross-sectional" = "#3366CC", "Longitudinal" = "#CC0000"),
      name = NULL
    ) +
    labs(
      title = unname(roi_labels[[modality]]),
      subtitle = plot_info$stats_label,
      x = "Age",
      y = paste("Neural", tolower(outcome)),
      caption = label
    ) +
    theme_classic(base_size = 10) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 6.8, hjust = 0.5, lineheight = 1.1),
      plot.caption = element_text(size = 8.5, hjust = 0.5),
      axis.title = element_text(size = 9.5),
      axis.text = element_text(size = 8.5),
      legend.text = element_text(size = 8.5)
    )

  list(
    figure = figure,
    roi_label = unname(roi_labels[[modality]]),
    stats_label = plot_info$stats_label
  )
}
```

```{r age-figure, fig.width=15, fig.height=5.2, out.width='100%'}
renderPlot({
  req(input$outcome)

  modalities <- c("Aud", "Mot", "Vis")
  panels <- lapply(
    modalities,
    function(modality) {
      make_roi_age_panel(
        analysis_data(),
        modality,
        input$outcome,
        ""
      )
    }
  )

  wrap_plots(lapply(panels, `[[`, "figure"), nrow = 1) +
    plot_layout(guides = "collect") +
    plot_annotation(
      title = paste(input$outcome, "age effects"),
      subtitle = rule_label(),
      theme = theme(
        plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 10)
      )
    ) &
    theme(legend.position = "bottom")
}, res = 110)
```

# Exploratory three-wave longitudinal analysis in MIND-C

This deliberately underpowered analysis asks whether the within-person trend
is also present in the MIND-C follow-up sample. Every included observation is
restricted to cognitively healthy participants (`Subgroup == 1`). MIND-C is
identified by the `mindc` Wave 3 subject ID, and matching Waves 1 and 2 are
linked using `SubNum`. A participant-modality pair must have retained data at
all three waves under the selected motion rule. The parsimonious exploratory
models adjust for modality and include a participant random intercept; no
multiplicity correction is applied.

```{r mindc-analysis, include=FALSE}
mindc_ids <- df_mind_long %>%
  filter(
    Subgroup == 1,
    Wave_Num == 3,
    stringr::str_starts(stringr::str_to_lower(Subject), "mindc")
  ) %>%
  distinct(SubNum) %>%
  pull(SubNum)

mindc_longitudinal_data <- reactive({
  motion_filtered_data() %>%
    filter(
      SubNum %in% mindc_ids,
      Subgroup == 1,
      Wave_Num %in% c(1, 2, 3)
    ) %>%
    group_by(SubNum, Modality) %>%
    filter(all(c(1, 2, 3) %in% Wave_Num)) %>%
    ungroup() %>%
    add_age_decomposition()
})

fit_mindc_model <- function(data, outcome) {
  model_data <- data %>%
    drop_na(
      all_of(c(
        outcome, "within_cAge", "aud_mot", "aud_vis", "SubNum"
      ))
    )

  validate(need(
    n_distinct(model_data$SubNum) >= 3,
    paste(
      "Fewer than three healthy MIND-C participants have complete",
      "three-wave data",
      "for", outcome, "under this motion rule."
    )
  ))

  lmer(
    reformulate(
      c("within_cAge", "aud_mot", "aud_vis", "(1 | SubNum)"),
      response = outcome
    ),
    data = model_data,
    REML = FALSE
  )
}

mindc_models <- reactive({
  data <- mindc_longitudinal_data()
  outcomes <- c("Reliability", "Confusability", "Distinctiveness")
  setNames(
    lapply(outcomes, function(outcome) {
      fit_mindc_model(data, outcome)
    }),
    outcomes
  )
})

mindc_longitudinal_effects <- reactive({
  map_dfr(names(mindc_models()), function(outcome) {
    model <- mindc_models()[[outcome]]
    coefficient <- summary(model)$coefficients[
      "within_cAge",
      ,
      drop = FALSE
    ]
    estimate <- coefficient[1, "Estimate"]
    standard_error <- coefficient[1, "Std. Error"]
    p_value <- coefficient[1, "Pr(>|t|)"]

    tibble(
      Outcome = outcome,
      B = estimate,
      SE = standard_error,
      `95% CI lower` = estimate - 1.96 * standard_error,
      `95% CI upper` = estimate + 1.96 * standard_error,
      t = coefficient[1, "t value"],
      df = coefficient[1, "df"],
      p = p_value,
      `Singular fit` = if_else(
        lme4::isSingular(model),
        "Yes",
        "No"
      )
    )
  })
})
```

## Available exploratory sample

```{r mindc-sample}
renderTable({
  data <- mindc_longitudinal_data()
  tibble(
    `Selected rule` = rule_label(),
    `Three-wave participants` = n_distinct(data$SubNum),
    `Complete three-wave participant-modality sets` =
      n_distinct(interaction(data$SubNum, data$Modality)),
    `Retained observations` = nrow(data)
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Estimated within-person age effects

`B` is the estimated change in the neural measure per one-year increase in
age. These estimates and *p* values should be treated as exploratory because
the three-wave MIND-C sample is very small. A singular-fit indicator is
included because the participant random-effect variance may not be estimable
reliably in such a small sample.

```{r mindc-effects-table}
renderTable({
  mindc_longitudinal_effects() %>%
    mutate(
      across(c(B, SE, `95% CI lower`, `95% CI upper`, t), ~ round(.x, 3)),
      df = round(df, 1),
      p = case_when(
        p < .001 ~ "< .001",
        TRUE ~ sprintf("%.3f", p)
      )
    )
}, striped = TRUE, bordered = TRUE, spacing = "s")
```

## Exploratory model summary

This display follows the outcome selected in the main outcome menu above.

```{r mindc-model-summary}
renderPrint({
  cat("Exploratory healthy MIND-C three-wave longitudinal model\n")
  cat(rule_label(), "\n")
  cat("Outcome:", input$outcome, "\n\n")
  print(summary(mindc_models()[[input$outcome]]))
})
```

## Three-wave trajectories

```{r mindc-trajectories, fig.width=13.5, fig.height=5.2, out.width='100%'}
renderPlot({
  plot_data <- mindc_longitudinal_data() %>%
    select(
      SubNum, Wave_Num, Age_DuringParticipation_, within_cAge, Modality,
      all_of(input$outcome)
    ) %>%
    rename(value = all_of(input$outcome)) %>%
    drop_na(value, Age_DuringParticipation_, within_cAge, SubNum)

  validate(need(
    nrow(plot_data) > 0,
    "No complete three-wave observations are available for this outcome."
  ))

  trajectory_stats <- plot_data %>%
    group_by(Modality) %>%
    group_modify(~ {
      modality_data <- .x
      within_sd <- sd(modality_data$within_cAge, na.rm = TRUE)

      if (n_distinct(modality_data$SubNum) < 3 ||
          !is.finite(within_sd) || within_sd == 0) {
        return(tibble(stats_label = "Longitudinal effect not estimable"))
      }

      tryCatch({
        modality_model <- lmer(
          value ~ within_cAge + (1 | SubNum),
          data = modality_data,
          REML = FALSE
        )
        coefficient <- summary(modality_model)$coefficients[
          "within_cAge",
          ,
          drop = FALSE
        ]
        estimate <- coefficient[1, "Estimate"]
        outcome_sd <- sd(modality_data$value, na.rm = TRUE)
        standardized_estimate <- if (is.finite(outcome_sd) && outcome_sd > 0) {
          estimate * within_sd / outcome_sd
        } else {
          NA_real_
        }

        tibble(
          stats_label = sprintf(
            "Longitudinal effect\nB = %.3f, β = %.3f\np = %s",
            estimate,
            standardized_estimate,
            format_age_p(coefficient[1, "Pr(>|t|)"])
          )
        )
      }, error = function(e) {
        tibble(stats_label = "Longitudinal effect not estimable")
      })
    }) %>%
    ungroup() %>%
    mutate(
      roi_label = recode(
        as.character(Modality),
        Aud = "Auditory",
        Mot = "Motor",
        Vis = "Visual"
      )
    )

  available_modalities <- c("Aud", "Mot", "Vis")[
    c("Aud", "Mot", "Vis") %in% as.character(unique(plot_data$Modality))
  ]

  roi_plots <- lapply(available_modalities, function(modality) {
    modality_data <- plot_data %>%
      filter(as.character(Modality) == modality)
    modality_stats <- trajectory_stats %>%
      filter(as.character(Modality) == modality)

    ggplot(
      modality_data,
      aes(
        x = Age_DuringParticipation_,
        y = value,
        group = SubNum
      )
    ) +
      geom_line(colour = "#7A7A7A", linewidth = 0.55, alpha = 0.7) +
      geom_point(aes(colour = factor(Wave_Num)), size = 2.5) +
      scale_y_continuous(expand = expansion(mult = c(0.08, 0.08))) +
      scale_colour_manual(
        values = c(
          "1" = "#3366CC",
          "2" = "#E68A00",
          "3" = "#CC0000"
        ),
        labels = c("Wave 1", "Wave 2", "Wave 3"),
        name = NULL
      ) +
      labs(
        title = modality_stats$roi_label[[1]],
        subtitle = modality_stats$stats_label[[1]],
        x = "Age",
        y = paste("Neural", tolower(input$outcome))
      ) +
      theme_classic(base_size = 11) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(
          face = "bold", size = 12, hjust = 0.5,
          margin = margin(b = 4)
        ),
        plot.subtitle = element_text(
          face = "plain", size = 8, hjust = 0.5, lineheight = 1.1,
          margin = margin(b = 8)
        ),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 9),
        legend.text = element_text(size = 9)
      )
  })

  trajectory_figure <- wrap_plots(roi_plots, nrow = 1) +
    plot_layout(guides = "collect") +
    plot_annotation(
      title = paste(
        "Exploratory healthy MIND-C three-wave trajectories:",
        input$outcome
      ),
      subtitle = rule_label(),
      theme = theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 9, hjust = 0.5)
      )
    ) &
    theme(legend.position = "bottom")

  trajectory_figure
}, res = 110)
```

# Excluded runs

```{r excluded-runs}
renderTable({
  selected_flags() %>%
    filter(exclude) %>%
    arrange(desc(pct_excluded), Subject, Modality) %>%
    select(Subject, Modality, pct_excluded) %>%
    rename(`Percent excluded TRs` = pct_excluded)
}, striped = TRUE, bordered = TRUE, spacing = "s")
```
